---
title: "Script_2.2_Prediction"
author: ""
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r libraries}
library(rstudioapi) #set WD
library(tidyverse) #%>% function
library(ggplot2)
library(caret) # machine learning
library(randomForest) # machine learning
library(reshape2) # For making long datasets (melt)
library(ranger)
library(patchwork)
```

```{r setting_WD_to_source_file_location}
setwd(dirname(getActiveDocumentContext()$path)) #setting Working directory
print(getwd()) # Checking working directory
```

```{r settings}
Variables_selected = 250
prediction_info_file_out = "results/total_ASV_prediction_info.txt"
total_prediction_VS_info_file_out = "results/total_ASV_VS_prediction_info.txt"
```

```{r load_data}
tab <- read.csv2("tab.txt", sep = "\t")
tax_tbl <- read.csv2("taxonomy_filtered.txt", sep = "\t")
metaData <- read.csv2("metaData.txt", sep = "\t")
```

```{r schaling}
read_counts<-rowSums(tab)
smallest<-min(read_counts)
normal<-smallest/read_counts
tab_rar<-tab*normal
```

```{r making_different_subsets}
# making the subsets based on metaData categories
metaData_fall <- metaData[!(metaData$timestamp == "P"), ]
metaData_summer <- metaData[(metaData$timestamp == "P"), ]
metaData_fall_mF <- metaData_fall[metaData_fall$sample_type == "soil", ]
metaData_fall_Time2 <- metaData_fall_mF[metaData_fall_mF$timestamp == "14", ]
metaData_fall_Time1 <- metaData_fall_mF[metaData_fall_mF$timestamp == "7", ]
metaData_fall_Time0 <- metaData_fall_mF[metaData_fall_mF$timestamp == "0", ]
metaData_fall_Time01 <- metaData_fall_mF[metaData_fall_mF$timestamp %in% c("7", "0"), ]
metaData_fall_Time12 <- metaData_fall_mF[metaData_fall_mF$timestamp %in% c("7", "14"), ]
metaData_summer_mF <- metaData_summer[metaData_summer$sample_type == "soil", ]
metaData_fall_fabric <- metaData_fall[metaData_fall$sample_type == "fabric", ]
metaData_summer_fabric <- metaData_summer[metaData_summer$sample_type == "fabric", ]
metaData_fall_summer <- metaData[metaData$place %in% metaData_summer$place , ]
metaData_fall_and_summer <- filter(metaData_fall_summer, place %in% c("A", "B", "C", "F"))
metaData_fall_ABCF <- filter(metaData_fall_mF, place %in% c("A", "B", "C", "F"))
metaData_summer_ABCF <- filter(metaData_summer_mF, place %in% c("A", "B", "C", "F"))

# making subsets of ASV tables based on metaData
tab_rar_fall <- (tab_rar[row.names(tab_rar) %in% row.names(metaData_fall), , drop = FALSE])
tab_rar_fall_mF <- (tab_rar[row.names(tab_rar) %in% row.names(metaData_fall_mF), , drop = FALSE])
tab_rar_summer <- (tab_rar[row.names(tab_rar) %in% row.names(metaData_summer), , drop = FALSE])
tab_rar_summer_mF <- (tab_rar[row.names(tab_rar) %in% row.names(metaData_summer_mF), , drop = FALSE])
tab_rar_fall_fabric <- (tab_rar[row.names(tab_rar) %in% row.names(metaData_fall_fabric), , drop = FALSE])
tab_rar_summer_fabric <- (tab_rar[row.names(tab_rar) %in% row.names(metaData_summer_fabric), , drop = FALSE])
tab_rar_fall_summer <- (tab_rar[row.names(tab_rar) %in% row.names(metaData_fall_summer), , drop = FALSE])
tab_rar_fall_Time2 <- (tab_rar[row.names(tab_rar) %in% row.names(metaData_fall_Time2), , drop = FALSE])
tab_rar_fall_Time1 <- (tab_rar[row.names(tab_rar) %in% row.names(metaData_fall_Time1), , drop = FALSE])
tab_rar_fall_Time0 <- (tab_rar[row.names(tab_rar) %in% row.names(metaData_fall_Time0), , drop = FALSE])
tab_rar_fall_Time01 <- (tab_rar[row.names(tab_rar) %in% row.names(metaData_fall_Time01), , drop = FALSE])
tab_rar_fall_Time12 <- (tab_rar[row.names(tab_rar) %in% row.names(metaData_fall_Time12), , drop = FALSE])
tab_rar_fall_ABCF <- (tab_rar_fall_mF[row.names(tab_rar_fall_mF) %in% row.names(metaData_fall_ABCF), , drop = FALSE])
tab_rar_summer_ABCF <- (tab_rar_summer_mF[row.names(tab_rar_summer_mF) %in% row.names(metaData_summer_ABCF), , drop = FALSE])

# removing ASVs not present in any of the samples in each table
keep<-colSums(tab_rar_fall)>0 
tab_rar_fall <-tab_rar_fall[,keep]
tax_fal <-tax_tbl[keep,]

keep<-colSums(tab_rar_fall_mF)>0 
tab_rar_fall_mF <-tab_rar_fall_mF[,keep]
tax_fall_mF <-tax_tbl[keep,]

keep<-colSums(tab_rar_summer)>0 
tab_rar_summer <-tab_rar_summer[,keep]
tax_summer <-tax_tbl[keep,]

keep<-colSums(tab_rar_summer_mF)>0 
table(keep)
tab_rar_summer_mF <-tab_rar_summer_mF[,keep]
tax_summer_mF <-tax_tbl[keep,]

keep<-colSums(tab_rar_summer_fabric)>0 
tab_rar_summer_fabric <-tab_rar_summer_fabric[,keep]
tax_summer_fabric <-tax_tbl[keep,]

keep<-colSums(tab_rar_fall_fabric)>0 
tab_rar_fall_fabric <-tab_rar_fall_fabric[,keep]
tax_fall_fabric <-tax_tbl[keep,]

keep<-colSums(tab_rar_fall_summer)>0 
tab_rar_fall_summer <-tab_rar_fall_summer[,keep]
tax_fall_summer <-tax_tbl[keep,]

keep<-colSums(tab_rar_fall_Time2)>0 
tab_rar_fall_Time2 <-tab_rar_fall_Time2[,keep]
tax_fall_Time2 <-tax_tbl[keep,]

keep<-colSums(tab_rar_fall_Time1)>0 
tab_rar_fall_Time1 <-tab_rar_fall_Time1[,keep]
tax_fall_Time1 <-tax_tbl[keep,]

keep<-colSums(tab_rar_fall_Time0)>0 
tab_rar_fall_Time0 <-tab_rar_fall_Time0[,keep]
tax_fall_Time0 <-tax_tbl[keep,]

keep<-colSums(tab_rar_fall_Time01)>0 
tab_rar_fall_Time01 <-tab_rar_fall_Time01[,keep]
tax_fall_Time01 <-tax_tbl[keep,]

keep<-colSums(tab_rar_fall_Time12)>0 
tab_rar_fall_Time12 <-tab_rar_fall_Time12[,keep]
tax_fall_Time12 <-tax_tbl[keep,]

keep<-colSums(tab_rar_fall_ABCF)>0 
tab_rar_fall_ABCF <-tab_rar_fall_ABCF[,keep]
tax_rar_fall_ABCF <-tax_tbl[keep,]

keep<-colSums(tab_rar_summer_ABCF)>0 
tab_rar_summer_ABCF <-tab_rar_summer_ABCF[,keep]
tax_rar_summer_ABCF <-tax_tbl[keep,]
keep <- NULL
```


#  Random forest using "Random Forest" package
```{r collect_pred_info}
prediction_info <- data.frame()
```

## Finding optimal mtry and ntree values
```{r optimizing, eval=FALSE}
X <- tab_rar
x_train <- tab_rar # Training features
y_train <- metaData$place # Training labels

# Define the search grid for hyperparameters including mtry, splitrule, and min.node.size
tune_grid <- expand.grid(
  mtry = c(floor(ncol(X)/6), floor(ncol(X)/5), floor(ncol(X)/4), floor(ncol(X)/3), floor(ncol(X)/2), floor(ncol(X)/1.5)),
  splitrule = "gini",  # Use the gini split rule which is commonly used for classification
  min.node.size = 1  # Setting a fixed min.node.size value
)
# Define control parameters for cross-validation
train_control <- trainControl(
  method = "cv",  # Use cross-validation
  number = 5,  # 5-fold cross-validation
  verboseIter = TRUE  # Print progress
)
# Train the Random Forest models for each combination of num.trees and mtry
results <- lapply(c(50, 100, 150, 200, 400, 600), function(num_trees) {
  set.seed(123)
  rf_cv_model <- train(
    x = x_train,
    y = y_train,
    method = "ranger",  # Use ranger for efficiency
    tuneGrid = tune_grid,  # The hyperparameter grid with mtry, splitrule, and min.node.size
    trControl = train_control,  # Cross-validation settings
    metric = "Accuracy",  # Metric to optimize
    num.trees = num_trees  # Specify number of trees 
  )
  # Store results along with the number of trees
  list(num.trees = num_trees, model = rf_cv_model, accuracy = max(rf_cv_model$results$Accuracy))
})

# Extract results into a dataframe
results_df <- do.call(rbind, lapply(results, function(res) {
  data.frame(num.trees = res$num.trees, mtry = res$model$bestTune$mtry, splitrule = res$model$bestTune$splitrule, min.node.size = res$model$bestTune$min.node.size, Accuracy = res$accuracy)
}))

# Find the optimal combination
best_result <- results_df[which.max(results_df$Accuracy), ]
print(best_result)
```



# Summer

## all summer samples
```{r prediction_of_summer_samples}
# Define feature matrix (X) and target vector (y)
X <- tab_rar_summer
y <- as.factor(metaData_summer$place)

# Create folds
folds_combined <- list()
for (i in 1:20) {
  set.seed(i)  # Set seed to the current value of i
  folds_combined <- c(folds_combined, createFolds(y, k = 4, list = TRUE, returnTrain = TRUE))
}


y_predTotal <- data.frame()
prob_list <- list()  
incorrect_predictions <- data.frame() 
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

for (i in 1:20){
  set.seed(2)
  train_index <- unlist(folds_combined[[i]])
  x_train <- X[train_index, ]
  x_test <- X[-train_index, ]
  y_train <- y[train_index]
  y_test <- y[-train_index]
  
  # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  
                           mtry = floor(ncol(x_train)/6),  
                           num.threads = parallel::detectCores() - 1, 
                           importance = TRUE)
# Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)
  
prob <- predict(rf_model, x_test, type = "prob") # finding probability scores
prob_list[[i]] <- prob  #saving the prob scores for each sample

#saving which samples were incorrectly predicted and what they were predicted as
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs here
      true_label = y_test[incorrect_idx], 
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
# Extract variable importance from the rf_model 
  importance_matrix <- as.data.frame(rf_model$importance) # Importance score for each ASV
  importance_matrix$ASV <-row.names(importance_matrix)
  
  # Sort by MeanDecreaseAccuracy and get top 
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  
  importance <- rbind(importance, top_importance)  # save the top scoring ASVs
  
  # Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[i] <- length(unique(noise$ASV))
unique_importance_count[i]<- length(unique(importance$ASV))
}

a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "E", "F", "G")
colnames(a) <- c("A", "B", "C", "E", "F", "G")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c/b

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")      
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'

# there are no incorrect predicted samples

```


```{r prediction_info}
always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

summer_prediction_info <- data.frame(
  Test = "summer all samples",
  Train = "summer all samples",
  Approach = "cross-validation",
  Accuracy = round(d, 3), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info , summer_prediction_info)
```

```{r confusion_matrix}
df <- as.data.frame(as.table(a))
# Calculate the row-wise sums
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))

# Plot the confusion matrix with custom axis titles
conf_matrix_fall_prediction_on_fabric <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 22),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") +
  ggtitle("B")

png("results/conf_matrix_ASV_summer_all.png", width = 1800, height = 1000)
print(conf_matrix_fall_prediction_on_fabric)
dev.off()
```

```{r probability_by_location}
prob_plot <- ggplot(prob_long, aes(x = predicted_place, y = Probability, color = place)) +
  geom_point(position = position_jitter(width = 0.1), size = 2) +
  facet_wrap(~ place, scales = "free_x") +
  labs(title = "Dot Plot of Probabilities by Sample and Class",
       x = "Sample",
       y = "Probability") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
    text = element_text(size = 30),  # Global text size
    axis.title = element_text(size = 30),  # Axis titles
    axis.text = element_text(size = 30),  # Axis text
    legend.title = element_text(size = 30),  # Legend title
    legend.text = element_text(size = 30)   # Legend text
  )
png("results/Probabilities_by_location_ASV_predicting_summer_plot.png", width = 1800, height = 1000)
print(prob_plot)
dev.off()
```

```{r Prob_distribution}
prob_long$Model <- rep("Summer", length(prob_long$Sample)) 
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]

# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]


# Add an identifier column to each data frame
prob_long_cor$Prediction <- "Correct"
prob_long_wrong$Prediction <- "Incorrect"

# Combine the data frames
prob_long_combined <- rbind(prob_long_cor, prob_long_wrong)

prob_distribution <- ggplot(prob_long_combined, aes(x = Probability, fill = Prediction)) +
  geom_histogram(alpha = 0.5, bins = 200) +  # Use alpha for transparency
  labs(title = "Distribution of Probabilities (Correct vs Incorrect Predictions)",
       x = "Probability",
       y = "Density") +
  scale_fill_manual(values = c("Correct" = "skyblue", "Incorrect" = "salmon")) +
  theme_minimal()
png("results/probability_distribution_summer.png", width = 1800, height = 1000)
print(prob_distribution)
dev.off()
```


```{r importance}
Importance_ASV_summer_all.tbl <- importance
```

## # Predicting with class taxonomy summer

```{r prediction_of_summer_samples}

tab_rar_summer_t <- as.data.frame(t(tab_rar_summer)) # Making samples the columns (so that tab rows and tax rows correspond)
abundance_summer_tab <- cbind(tax_summer, tab_rar_summer_t)

filtered_abundance_tab <- dplyr::select(abundance_summer_tab, !c(Family, Order, Kingdom, Phylum, Genus, Species))

  # Summarize numeric columns by the "Class" column
  summarized_data <- filtered_abundance_tab %>%
    group_by(Class) %>%
    summarize(across(where(is.numeric), sum, na.rm = TRUE))

summarized_data <- as.data.frame(filter(summarized_data, Class != "NA"))
row.names(summarized_data) <- summarized_data$Class
summarized_data <- dplyr::select(summarized_data, !Class)
summarized_data_t <- as.data.frame(t(summarized_data))

# Define feature matrix (X) and target vector (y)
X <- summarized_data_t
y <- as.factor(metaData_summer$place)



folds_combined <- list()
for (i in 1:20) {
  set.seed(i)  # Set seed to the current value of i
  folds_combined <- c(folds_combined, createFolds(y, k = 4, list = TRUE, returnTrain = TRUE))
}


y_predTotal <- data.frame()
prob_list <- list()  
incorrect_predictions <- data.frame() 
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

for (i in 1:20){
  set.seed(2)
  train_index <- unlist(folds_combined[[i]])
  x_train <- X[train_index, ]
  x_test <- X[-train_index, ]
  y_train <- y[train_index]
  y_test <- y[-train_index]
  
  # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  
                           mtry = floor(ncol(x_train)/6),  
                           num.threads = parallel::detectCores() - 1, 
                           importance = TRUE)
# Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)
  
prob <- predict(rf_model, x_test, type = "prob") # finding probability scores
prob_list[[i]] <- prob  #saving the prob scores for each sample

#saving which samples were incorrectly predicted and what they were predicted as
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs here
      true_label = y_test[incorrect_idx], 
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
# Extract variable importance from the rf_model 
  importance_matrix <- as.data.frame(rf_model$importance) # Importance score for each ASV
  importance_matrix$ASV <-row.names(importance_matrix)
  
  # Sort by MeanDecreaseAccuracy and get top 
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  
  importance <- rbind(importance, top_importance)  # save the top scoring ASVs
  
  # Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[i] <- length(unique(noise$ASV))
unique_importance_count[i]<- length(unique(importance$ASV))
}

a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "E", "F", "G")
colnames(a) <- c("A", "B", "C", "E", "F", "G")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c/b

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")      
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'


# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%  # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe


```

```{r prediction_info}
always_incorrect_ids <-  paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")


summer_prediction_with_taxa_info <- data.frame(
  Test = "summer samples",
  Train = "summer samples",
  Approach = "cross-validation, class taxa as features",
  Accuracy = round(d, 4), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info , summer_prediction_with_taxa_info)
```

```{r confusion_matrix}
df <- as.data.frame(as.table(a))
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))

# Plot the confusion matrix with custom axis titles
conf_matrix_fall_prediction_on_fabric <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 22),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") +
  ggtitle("B")

png("results/conf_matrix_ASV_summer_taxa.png", width = 1800, height = 1000)
print(conf_matrix_fall_prediction_on_fabric)
dev.off()
```


```{r importance}
Importance_ASV_summer_With_taxa.tbl <- importance

```

# Fall

## soil from fall


```{r prediction_of_fall_soil_samples}

# Define feature matrix (X) and target vector (y)
X <- tab_rar_fall_mF
y <- as.factor(metaData_fall_mF$place)


folds_combined <- list()
for (i in 1:20) {
  set.seed(i)  # Set seed to the current value of i
  folds_combined <- c(folds_combined, createFolds(y, k = 5, list = TRUE, returnTrain = TRUE))
}

# Containers needed for collecting prediction info
y_predTotal <- data.frame()
prob_list <- list()  
incorrect_predictions <- data.frame() 
importance <- data.frame()  
noise <- data.frame()  
unique_noise_count <- numeric()
unique_importance_count <- numeric()

for (i in 1:25){ # Each sample is predicted 5 times
  set.seed(2)
  train_index <- unlist(folds_combined[[i]]) 
  x_train <- X[train_index, ]
  x_test <- X[-train_index, ]
  y_train <- y[train_index]
  y_test <- y[-train_index]
  
  # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  
                           mtry = floor(ncol(x_train)/6),  
                           num.threads = parallel::detectCores() - 1, 
                           importance = TRUE)
# Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)
  
prob <- predict(rf_model, x_test, type = "prob") # finding probability scores
prob_list[[i]] <- prob  #saving the prob scores for each sample

#saving which samples were incorrectly predicted and what they were predicted as
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs here
      true_label = y_test[incorrect_idx], 
      predicted_label = y_pred[incorrect_idx],
      Fold = i
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
# Extract variable importance from the rf_model 
  importance_matrix <- as.data.frame(rf_model$importance) # Importance score for each ASV
  importance_matrix$ASV <-row.names(importance_matrix)
  
  # Sort by MeanDecreaseAccuracy and get top 
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  
  importance <- rbind(importance, top_importance)  # save the top scoring ASVs
  
  # Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[i] <- length(unique(noise$ASV))
unique_importance_count[i]<- length(unique(importance$ASV))
}

a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "L", "N", "P", "Q", "R")
colnames(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "L", "N", "P", "Q", "R")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c/b

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")  
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'

# inspecting incorrectly predicted samples
incorrect_pred <- incorrect_predictions
incorrect_predictions <- incorrect_pred %>%
  group_by(sample_id) %>%  
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe

```

```{r prediction_info}
always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

fall_soil_prediction_info <- data.frame(
  Test = "fall soil",
  Train = "fall soil",
  Approach = "cross-validation",
  Accuracy = round(d, 3), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)

prediction_info <- rbind(prediction_info , fall_soil_prediction_info)
```

```{r formating_confusion_matrix}
df <- as.data.frame(as.table(a))
# Calculate the row-wise sums
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix_soil_fall <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 30),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_soil_fall.png", width = 1800, height = 1000)
print(conf_matrix_soil_fall)
dev.off()
```

```{r Prob_distribution}
prob_long$Model <- rep("Soil_fall", length(prob_long$Sample))
prob_long_cor1 <- prob_long[prob_long$place == prob_long$predicted_place, ]

# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'
prob_long_wrong1 <- prob_long[prob_long$place != prob_long$predicted_place, ]


# Add an identifier column to each data frame
prob_long_cor1$Prediction <- "Correct"
prob_long_wrong1$Prediction <- "Incorrect"

# Combine the data frames
prob_long_combined <- rbind(prob_long_cor1, prob_long_wrong1)

prob_distribution <- ggplot(prob_long_combined, aes(x = Probability, fill = Prediction)) +
  geom_histogram(alpha = 0.5, bins = 200) +  # Use alpha for transparency
  labs(title = "Distribution of Probabilities (Correct vs Incorrect Predictions)",
       x = "Probability",
       y = "Density") +
  scale_fill_manual(values = c("Correct" = "skyblue", "Incorrect" = "salmon")) +
  theme_minimal()
png("results/probability_distribution_ASV_fall_soil.png", width = 1800, height = 1000)
print(prob_distribution)
dev.off()
```



```{r Prob_distribution_by_location}
prob_plot <- ggplot(prob_long, aes(x = predicted_place, y = Probability, color = place)) +
  geom_point(position = position_jitter(width = 0.1), size = 2) +
  facet_wrap(~ place, scales = "free_x") +
  labs(title = "Dot Plot of Probabilities by Sample and Class",
       x = "Sample",
       y = "Probability") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
    text = element_text(size = 20),  # Global text size
    axis.title = element_text(size = 20),  # Axis titles
    axis.text = element_text(size = 20),  # Axis text
    legend.title = element_text(size = 20),  # Legend title
    legend.text = element_text(size = 20)   # Legend text
  )
png("results/Probabilities_by_location_ASV_fall_soil_plot.png", width = 1800, height = 1000)
print(prob_plot)
dev.off()
```

```{r importance_ASVs}
importance_ASV_fall_soil_samples.tbl <- importance
```


## Soil fall with class taxonomy

```{r prediction_of_fall_soil_samples}
tab_rar_fall_t <- as.data.frame(t(tab_rar_fall_mF)) # Making samples the columns (so that tab rows and tax rows correspond)
abundance_fall_tab <- cbind(tax_fall_mF, tab_rar_fall_t)

filtered_abundance_tab <- dplyr::select(abundance_fall_tab, !c(Family, Order, Kingdom, Phylum, Genus, Species))

  # Summarize numeric columns by the "Class" column
  summarized_data <- filtered_abundance_tab %>%
    group_by(Class) %>%
    summarize(across(where(is.numeric), sum, na.rm = TRUE))

summarized_data <- as.data.frame(filter(summarized_data, Class != "NA"))
row.names(summarized_data) <- summarized_data$Class
summarized_data <- dplyr::select(summarized_data, !Class)
summarized_data_t <- as.data.frame(t(summarized_data))

# Define feature matrix (X) and target vector (y)
X <- summarized_data_t
y <- as.factor(metaData_fall_mF$place)

folds_combined <- list()
for (i in 1:20) {
  set.seed(i)  # Set seed to the current value of i
  folds_combined <- c(folds_combined, createFolds(y, k = 5, list = TRUE, returnTrain = TRUE))
}



y_predTotal <- data.frame()
prob_list <- list()  
incorrect_predictions <- data.frame() 
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

for (i in 1:25){
  set.seed(2)
  train_index <- unlist(folds_combined[[i]]) 
  x_train <- X[train_index, ]
  x_test <- X[-train_index, ]
  y_train <- y[train_index]
  y_test <- y[-train_index]
  
  # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  
                           mtry = floor(ncol(x_train)/6),  
                           num.threads = parallel::detectCores() - 1, 
                           importance = TRUE)
# Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)
  
prob <- predict(rf_model, x_test, type = "prob") # finding probability scores
prob_list[[i]] <- prob  #saving the prob scores for each sample

#saving which samples were incorrectly predicted and what they were predicted as
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs here
      true_label = y_test[incorrect_idx], 
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
# Extract variable importance from the rf_model 
  importance_matrix <- as.data.frame(rf_model$importance) # Importance score for each ASV
  importance_matrix$ASV <-row.names(importance_matrix)
  
  # Sort by MeanDecreaseAccuracy and get top 
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  
  importance <- rbind(importance, top_importance)  # save the top scoring ASVs
  
  # Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[i] <- length(unique(noise$ASV))
unique_importance_count[i]<- length(unique(importance$ASV))
}

a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "L", "N", "P", "Q", "R")
colnames(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "L", "N", "P", "Q", "R")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c/b

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")     
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'

# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%  # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe

```

```{r prediction_info}
always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

Class_prediction_info <- data.frame(
  Test = "soil fall",
  Train = "soil fall",
  Approach = "cross-validation, class taxa as features",
   Accuracy = round(d, 4),  
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info, Class_prediction_info)
```

```{r conf_matrix}
df <- as.data.frame(as.table(a))
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 30),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_fall_prediction_with_taxonomy_at_class_level.png", width = 1800, height = 1000)
print(conf_matrix)
dev.off()
```


```{r importance}
Importance_ASV_fall_class_prediction.tbl <- importance
```


## prediction of soil from fall with variables selected as important during fall soil prediction -finding highest accuracy
```{r}
# sorting important ASVs based on frequency (for how many models they are amongst the 250 most important)
asv_freq <- as.data.frame(table(importance_ASV_fall_soil_samples.tbl$ASV))
sorted_asv_freq <- asv_freq[order(-asv_freq$Freq), ]

# Initialize a dataframe to store results
VS_info <- data.frame()

for (n_top_asvs in seq(10, 500, by = 10)) {
  
  # Select top n ASVs
  top_ASVs <- head(sorted_asv_freq, n_top_asvs)
  selected_ASVs <- as.character(top_ASVs$Var1)
  
  # Subset the feature table
  tab_rar_fall_mF_subset <- tab_rar_fall_mF[, colnames(tab_rar_fall_mF) %in% selected_ASVs]
  
  # Define X and y
  X <- tab_rar_fall_mF_subset
  y <- as.factor(metaData_fall_mF$place)
  
  # Create folds
  folds_combined <- list()
  for (i in 1:20) {
    set.seed(i)
    folds_combined <- c(folds_combined, createFolds(y, k = 5, list = TRUE, returnTrain = TRUE))
  }

  # Initialize containers
  y_predTotal <- data.frame()
  importance <- data.frame()
  noise <- data.frame()
  
    Variables_selected <- n_top_asvs
  
  for (i in 1:25) {
    set.seed(2)
    train_index <- unlist(folds_combined[[i]])
    x_train <- X[train_index, ]
    x_test <- X[-train_index, ]
    y_train <- y[train_index]
    y_test <- y[-train_index]
    
    rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,
                             mtry = floor(ncol(x_train)/6),
                             num.threads = parallel::detectCores() - 1,
                             importance = TRUE)
    
    y_pred <- predict(rf_model, x_test)
    y_predRes <- data.frame(y_pred = y_pred, y_test = y_test)
    y_predTotal <- rbind(y_predTotal, y_predRes)
    
   
    importance_matrix <- as.data.frame(rf_model$importance)
    importance_matrix$ASV <- rownames(importance_matrix)
    top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]
    top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]
    importance <- rbind(importance, top_importance)
    
    noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
    noise <- rbind(noise, noise_vars)
    
    }
  
  b <- nrow(y_predTotal)
  c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
  d <- c / b
  
   fall_soil_nrASVs_VS_info <- data.frame(
    Model = paste0("Top_", n_top_asvs, "_ASVs"),
    Accuracy = d,
    Features = ncol(X))
  
  VS_info <- rbind(VS_info, fall_soil_nrASVs_VS_info)
}

```


```{r}
png("results/nr_important_fall_soil_ASVs_for_soil_fall_pred", width = 1800, height = 1000)
# Create the scatter plot
ggplot(VS_info, aes(x = Features, y = Accuracy)) +
  geom_point(size = 3, color = "steelblue") +
  geom_line(color = "steelblue", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Model Accuracy vs Number of Features",
    x = "Number of Features (ASVs)",
    y = "Prediction Accuracy"
  )
dev.off()

```


## prediction of soil from fall with nr of important variables giving highest accuracy 
```{r}
top_ASVs <- head(sorted_asv_freq,  VS_info$Features[which.max(VS_info$Accuracy)])
selected_ASVs <- as.character(top_ASVs$Var1)
selected_ASVs_fall_soil <- selected_ASVs
```

```{r prediction_of_fall_soil_samples}
# Subset the feature table
  tab_rar_fall_mF_subset <- tab_rar_fall_mF[, colnames(tab_rar_fall_mF) %in% selected_ASVs]
  
  # Define X and y
  X <- tab_rar_fall_mF_subset
  y <- as.factor(metaData_fall_mF$place)
  
  folds_combined <- list()
for (i in 1:20) {
  set.seed(i)  # Set seed to the current value of i
  folds_combined <- c(folds_combined, createFolds(y, k = 5, list = TRUE, returnTrain = TRUE))
}


# Containers needed for collecting prediction info
y_predTotal <- data.frame()
prob_list <- list()  
incorrect_predictions <- data.frame() 
importance <- data.frame()  
noise <- data.frame()  
unique_noise_count <- numeric()
unique_importance_count <- numeric()

for (i in 1:25){ # Each sample is predicted 5 times
  set.seed(2)
  train_index <- unlist(folds_combined[[i]]) 
  x_train <- X[train_index, ]
  x_test <- X[-train_index, ]
  y_train <- y[train_index]
  y_test <- y[-train_index]
  
  # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  
                           mtry = floor(ncol(x_train)/6),  
                           num.threads = parallel::detectCores() - 1, 
                           importance = TRUE)
# Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)
  
prob <- predict(rf_model, x_test, type = "prob") # finding probability scores
prob_list[[i]] <- prob  #saving the prob scores for each sample

#saving which samples were incorrectly predicted and what they were predicted as
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs here
      true_label = y_test[incorrect_idx], 
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
# Extract variable importance from the rf_model 
  importance_matrix <- as.data.frame(rf_model$importance) # Importance score for each ASV
  importance_matrix$ASV <-row.names(importance_matrix)
  
  # Sort by MeanDecreaseAccuracy and get top 
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  
  importance <- rbind(importance, top_importance)  # save the top scoring ASVs
  
  # Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[i] <- length(unique(noise$ASV))
unique_importance_count[i]<- length(unique(importance$ASV))
}

a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "L", "N", "P", "Q", "R")
colnames(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "L", "N", "P", "Q", "R")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c/b

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")  
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'

# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%  # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe
```

```{r prediction_info}
always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

fall_soil_prediction_VS_info <- data.frame(
  Test = "fall soil",
  Train = "fall soil",
  Approach = "cross-validation, Feature selection",
  Accuracy = round(d, 4),  
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)

prediction_info <- rbind(prediction_info, fall_soil_prediction_VS_info)
```

```{r formating_confusion_matrix}
df <- as.data.frame(as.table(a))

row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix_soil_fall <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 30),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_soil_fall_VS.png", width = 1800, height = 1000)
print(conf_matrix_soil_fall)
dev.off()
```

```{r Prob_distribution}
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]

# Add an identifier column to each data frame
prob_long_cor$Prediction <- "Correct"
prob_long_wrong$Prediction <- "Incorrect"

# Combine the data frames
prob_long_combined <- rbind(prob_long_cor, prob_long_wrong)

prob_distribution <- ggplot(prob_long_combined, aes(x = Probability, fill = Prediction)) +
  geom_density(alpha = 0.5, bins = 200) +  # Use alpha for transparency
  labs(title = "Distribution of Probabilities (Correct vs Incorrect Predictions)",
       x = "Probability",
       y = "Density") +
  scale_fill_manual(values = c("Correct" = "skyblue", "Incorrect" = "salmon")) +
  theme_minimal()
png("results/probability_distribution_ASV_fall_soil_VS.png", width = 1800, height = 1000)
print(prob_distribution)
dev.off()
```

```{r Prob_distribution_by_location}
prob_plot <- ggplot(prob_long, aes(x = predicted_place, y = Probability, color = place)) +
  geom_point(position = position_jitter(width = 0.1), size = 2) +
  facet_wrap(~ place, scales = "free_x") +
  labs(title = "Dot Plot of Probabilities by Sample and Class",
       x = "Sample",
       y = "Probability") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
    text = element_text(size = 20),  # Global text size
    axis.title = element_text(size = 20),  # Axis titles
    axis.text = element_text(size = 20),  # Axis text
    legend.title = element_text(size = 20),  # Legend title
    legend.text = element_text(size = 20)   # Legend text
  )
png("results/Probabilities_by_location_ASV_fall_soil_plot_VS.png", width = 1800, height = 1000)
print(prob_plot)
dev.off()
```

# Sampling day

## Predicting week 2 when training with the other timepoints

```{r}
Place_Time2 <- unique(metaData_fall_Time2$place)
metaData_fall_Time01_subset <- dplyr::filter(metaData_fall_Time01, place %in% Place_Time2)
tab_rar_fall_Time01_subset <- (tab_rar_fall_Time01[row.names(tab_rar_fall_Time01) %in% row.names(metaData_fall_Time01_subset), , drop = FALSE])

keep<-colSums(tab_rar_fall_Time01_subset)>0 
tab_rar_fall_Time01_subset <-tab_rar_fall_Time01_subset[,keep]
common_columns <- intersect(colnames(tab_rar_fall_Time01_subset), colnames(tab_rar_fall_Time2))

# Subset both tables to only the common columns
tab_rar_fall_Time01_subset <- tab_rar_fall_Time01_subset[, common_columns]
tab_rar_fall_Time2_subset <- tab_rar_fall_Time2[, common_columns]

# Train and test data
x_train <- tab_rar_fall_Time01_subset  # feature matrix for training
y_train <- as.factor(metaData_fall_Time01_subset$place)  # target for training
x_test <- tab_rar_fall_Time2_subset  # feature matrix for testing
y_test <- as.factor(metaData_fall_Time2$place)  # target for testing (assuming first letter of row names)

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

for (seed in 1:5) {
  set.seed(seed)  # Set the seed
  
      # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  # Can adjust num.trees
                           mtry = floor(ncol(x_train)/6),  # Number of variables to consider at each split
                           importance = TRUE, num.threads = parallel::detectCores() - 1)
  
  # Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)

  # Get probabilities for the test set
  prob <- predict(rf_model, x_test, type = "prob")
  prob_list[[seed]] <- prob

  # Collect incorrect predictions
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs
      true_label = y_test[incorrect_idx],
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
  
  # Extract variable importance from the rf_model object directly
  importance_matrix <- as.data.frame(rf_model$importance)
  importance_matrix$ASV <- rownames(importance_matrix)  # Add ASV names
  importance_matrix$seed <- seed  # Add seed value
  
# Sort by MeanDecreaseAccuracy and get top
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  # Most informative variables
  importance <- rbind(importance, top_importance)  # Append to the importance table
  
# Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[seed] <- length(unique(noise$ASV))
unique_importance_count[seed]<- length(unique(importance$ASV))
}

# Summarize predictions
a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c(unique(metaData_fall_Time01_subset$place))
colnames(a) <- c(unique(metaData_fall_Time01_subset$place))
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c / b

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")     
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'


# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%  # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe

```

```{r prediction_info}
always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

fall_Time2_prediction_info <- data.frame(
  Test = "fall day 14 soil",
  Train = "fall day 0 and 7 soil",
  Approach = " ",
  Accuracy = round(d, 4), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info , fall_Time2_prediction_info)
```

```{r confusion_matrix}
df <- as.data.frame(as.table(a))
Var1 <- unique(metaData_fall_Time0$place)
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix_fall_prediction_on_fabric <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 22),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_fall_prediction_on_timepoint14.png", width = 1800, height = 1000)
print(conf_matrix_fall_prediction_on_fabric)
dev.off()
```


```{r Prob_distribution}
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]

# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]

# Add an identifier column to each data frame
prob_long_cor$Prediction <- "Correct"
prob_long_wrong$Prediction <- "Incorrect"

# Combine the data frames
prob_long_combined <- rbind(prob_long_cor, prob_long_wrong)

prob_distribution <- ggplot(prob_long_combined, aes(x = Probability, fill = Prediction)) +
  geom_histogram(alpha = 0.5, bins = 200) +  # Use alpha for transparency
  labs(title = "Distribution of Probabilities (Correct vs Incorrect Predictions)",
       x = "Probability",
       y = "Density") +
  scale_fill_manual(values = c("Correct" = "skyblue", "Incorrect" = "salmon")) +
  theme_minimal()
png("results/probability_distribution_week2.png", width = 1800, height = 1000)
print(prob_distribution)
dev.off()
```

```{r Prob_distribution_by_location}
prob_plot <- ggplot(prob_long, aes(x = predicted_place, y = Probability, color = place)) +
  geom_point(position = position_jitter(width = 0.1), size = 2) +
  facet_wrap(~ place, scales = "free_x") +
  labs(title = "Dot Plot of Probabilities by Sample and Class",
       x = "Sample",
       y = "Probability") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
    text = element_text(size = 20),  # Global text size
    axis.title = element_text(size = 20),  # Axis titles
    axis.text = element_text(size = 20),  # Axis text
    legend.title = element_text(size = 20),  # Legend title
    legend.text = element_text(size = 20)   # Legend text
  )
png("results/Probabilities_by_location_ASV_fall_Time14_plot.png", width = 1800, height = 1000)
print(prob_plot)
dev.off()
```


```{r unique_importance}
Importance_ASV_fall_Time2.tbl <- importance
```

## Predicting week 2 when training with the other timepoints precense absence data
```{r}

tab_rar_fall_Time01_subset[tab_rar_fall_Time01_subset > 0] <- 1
tab_rar_fall_Time2_subset[tab_rar_fall_Time2_subset > 0] <- 1

# Train and test data
x_train <- tab_rar_fall_Time01_subset  # feature matrix for training
y_train <- as.factor(metaData_fall_Time01_subset$place)  # target for training
x_test <- tab_rar_fall_Time2_subset  # feature matrix for testing
y_test <- as.factor(metaData_fall_Time2$place)  # target for testing (assuming first letter of row names)

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

for (seed in 1:5) {
  set.seed(seed)  # Set the seed
  
  # Train the Random Forest model
    # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  # Can adjust num.trees
                           mtry = floor(ncol(x_train)/6),  # Number of variables to consider at each split
                           importance = TRUE, num.threads = parallel::detectCores() - 1)
  
  # Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)

  # Get probabilities for the test set
  prob <- predict(rf_model, x_test, type = "prob")
  prob_list[[seed]] <- prob

  # Collect incorrect predictions
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs
      true_label = y_test[incorrect_idx],
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
  
  # Extract variable importance from the rf_model object directly
  importance_matrix <- as.data.frame(rf_model$importance)
  importance_matrix$ASV <- rownames(importance_matrix)  # Add ASV names
  importance_matrix$seed <- seed  # Add seed value
  
# Sort by MeanDecreaseAccuracy and get top
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  # Most informative variables
  importance <- rbind(importance, top_importance)  # Append to the importance table
  
# Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[seed] <- length(unique(noise$ASV))
unique_importance_count[seed]<- length(unique(importance$ASV))
}

# Summarize predictions
a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c(unique(metaData_fall_Time01_subset$place))
colnames(a) <- c(unique(metaData_fall_Time01_subset$place))
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c / b

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")     
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'


# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%  # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe
```

```{r prediction_info}
always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")


fall_Time2_prediction_pa_info <- data.frame(
  Test = "fall day 14 soil",
  Train = "fall day 0 and 7 soil",
  Approach = "presence/absense",
  Accuracy = round(d, 4),  
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info , fall_Time2_prediction_pa_info)
```

```{r confusion_matrix}
df <- as.data.frame(as.table(a))
Var1 <- unique(metaData_fall_Time0$place)
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix_fall_prediction_on_fabric <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 22),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_fall_prediction_on_timepoint14_presence_absence.png", width = 1800, height = 1000)
print(conf_matrix_fall_prediction_on_fabric)
dev.off()
```


```{r Prob_distribution}
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]

# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]

# Add an identifier column to each data frame
prob_long_cor$Prediction <- "Correct"
prob_long_wrong$Prediction <- "Incorrect"

# Combine the data frames
prob_long_combined <- rbind(prob_long_cor, prob_long_wrong)

prob_distribution <- ggplot(prob_long_combined, aes(x = Probability, fill = Prediction)) +
  geom_histogram(alpha = 0.5, bins = 200) +  # Use alpha for transparency
  labs(title = "Distribution of Probabilities (Correct vs Incorrect Predictions)",
       x = "Probability",
       y = "Density") +
  scale_fill_manual(values = c("Correct" = "skyblue", "Incorrect" = "salmon")) +
  theme_minimal()
png("results/probability_distribution_week2_precence_absense.png", width = 1800, height = 1000)
print(prob_distribution)
dev.off()
```

```{r Prob_distribution_by_location}
prob_plot <- ggplot(prob_long, aes(x = predicted_place, y = Probability, color = place)) +
  geom_point(position = position_jitter(width = 0.1), size = 2) +
  facet_wrap(~ place, scales = "free_x") +
  labs(title = "Dot Plot of Probabilities by Sample and Class",
       x = "Sample",
       y = "Probability") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
    text = element_text(size = 20),  # Global text size
    axis.title = element_text(size = 20),  # Axis titles
    axis.text = element_text(size = 20),  # Axis text
    legend.title = element_text(size = 20),  # Legend title
    legend.text = element_text(size = 20)   # Legend text
  )
png("results/Probabilities_by_location_ASV_fall_week2_precence_absense_plot.png", width = 1800, height = 1000)
print(prob_plot)
dev.off()
```


```{r unique_importance}
Importance_ASV_fall_Time2_precence_absense.tbl <- importance
```


## Predicting week 0 when training with the other timepoints

```{r}
common_columns <- intersect(colnames(tab_rar_fall_Time12), colnames(tab_rar_fall_Time0))

# Subset both tables to only the common columns
tab_rar_fall_Time12_subset <- tab_rar_fall_Time12[, common_columns]
tab_rar_fall_Time0_subset <- tab_rar_fall_Time0[, common_columns]

# Train and test data
x_train <- tab_rar_fall_Time12_subset  # feature matrix for training
y_train <- as.factor(metaData_fall_Time12$place)  # target for training
x_test <- tab_rar_fall_Time0_subset  # feature matrix for testing
y_test <- as.factor(metaData_fall_Time0$place)  # target for testing (assuming first letter of row names)

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

for (seed in 1:5) {
  set.seed(seed)  # Set the seed
  
  # Train the Random Forest model
    # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  # Can adjust num.trees
                           mtry = floor(ncol(x_train)/6),  # Number of variables to consider at each split
                           importance = TRUE, num.threads = parallel::detectCores() - 1)
  
  # Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)

  # Get probabilities for the test set
  prob <- predict(rf_model, x_test, type = "prob")
  prob_list[[seed]] <- prob

  # Collect incorrect predictions
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs
      true_label = y_test[incorrect_idx],
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
  
  # Extract variable importance from the rf_model object directly
  importance_matrix <- as.data.frame(rf_model$importance)
  importance_matrix$ASV <- rownames(importance_matrix)  # Add ASV names
  importance_matrix$seed <- seed  # Add seed value
  
# Sort by MeanDecreaseAccuracy and get top
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  # Most informative variables
  importance <- rbind(importance, top_importance)  # Append to the importance table
  
# Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[seed] <- length(unique(noise$ASV))
unique_importance_count[seed]<- length(unique(importance$ASV))
}

# Summarize predictions
a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "L", "N", "P", "Q", "R")
colnames(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "N", "P", "Q", "R")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c / b

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")      
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'

# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%  # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe

```

```{r prediction_info}
always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

fall_Time0_prediction_info <- data.frame(
  Test = "fall day 0 soil",
  Train = "fall day 7 and 14 soil",
  Approach = " ",
  Accuracy = round(d, 4),  
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info , fall_Time0_prediction_info)
```

```{r confusion_matrix}
df <- as.data.frame(as.table(a))
Var1 <- unique(metaData_fall_Time0$place)

# Create a data frame with Var1, Var2 filled with "Q", and Freq filled with 0
df2 <- data.frame(
  Var1 = Var1,             # First column
  Var2 = rep("L", length(Var1)),  # Second column with "Q"
  Freq = rep(0, length(Var1))     # Third column with 0s
)
df <- rbind(df, df2)
df$Var2 <- as.character(df$Var2)
idx <- order(df$Var2)
df <- df[idx,]
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix_fall_prediction_on_fabric <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 22),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_fall_prediction_on_timepoint0.png", width = 1800, height = 1000)
print(conf_matrix_fall_prediction_on_fabric)
dev.off()
```

```{r Prob_distribution_by_location}
prob_plot <- ggplot(prob_long, aes(x = predicted_place, y = Probability, color = place)) +
  geom_point(position = position_jitter(width = 0.1), size = 2) +
  facet_wrap(~ place, scales = "free_x") +
  labs(title = "Dot Plot of Probabilities by Sample and Class",
       x = "Sample",
       y = "Probability") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
    text = element_text(size = 20),  # Global text size
    axis.title = element_text(size = 20),  # Axis titles
    axis.text = element_text(size = 20),  # Axis text
    legend.title = element_text(size = 20),  # Legend title
    legend.text = element_text(size = 20)   # Legend text
  )
png("results/Probabilities_by_location_ASV_fall_time0.png", width = 1800, height = 1000)
print(prob_plot)
dev.off()
```


```{r Prob_distribution}
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]

# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]


# Add an identifier column to each data frame
prob_long_cor$Prediction <- "Correct"
prob_long_wrong$Prediction <- "Incorrect"

# Combine the data frames
prob_long_combined <- rbind(prob_long_cor, prob_long_wrong)

prob_distribution <- ggplot(prob_long_combined, aes(x = Probability, fill = Prediction)) +
  geom_histogram(alpha = 0.5, bins = 200) +  # Use alpha for transparency
  labs(title = "Distribution of Probabilities (Correct vs Incorrect Predictions)",
       x = "Probability",
       y = "Density") +
  scale_fill_manual(values = c("Correct" = "skyblue", "Incorrect" = "salmon")) +
  theme_minimal()
png("results/probability_distribution_week0.png", width = 1800, height = 1000)
print(prob_distribution)
dev.off()
```


```{r unique_importance}
Importance_ASV_fall_Time0.tbl <- importance

```


## Predicting week 0 when training with the other timepoints presence absence

```{r}
tab_rar_fall_Time12_subset[tab_rar_fall_Time12_subset > 0] <- 1
tab_rar_fall_Time0_subset[tab_rar_fall_Time0_subset > 0] <- 1

# Train and test data
x_train <- tab_rar_fall_Time12_subset  # feature matrix for training
y_train <- as.factor(metaData_fall_Time12$place)  # target for training
x_test <- tab_rar_fall_Time0_subset  # feature matrix for testing
y_test <- as.factor(metaData_fall_Time0$place)  # target for testing (assuming first letter of row names)

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

for (seed in 1:5) {
  set.seed(seed)  # Set the seed
  
  # Train the Random Forest model
    # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  # Can adjust num.trees
                           mtry = floor(ncol(x_train)/6),  # Number of variables to consider at each split
                           importance = TRUE, num.threads = parallel::detectCores() - 1)
  
  # Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)

  # Get probabilities for the test set
  prob <- predict(rf_model, x_test, type = "prob")
  prob_list[[seed]] <- prob

  # Collect incorrect predictions
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs
      true_label = y_test[incorrect_idx],
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
  
  # Extract variable importance from the rf_model object directly
  importance_matrix <- as.data.frame(rf_model$importance)
  importance_matrix$ASV <- rownames(importance_matrix)  # Add ASV names
  importance_matrix$seed <- seed  # Add seed value
  
# Sort by MeanDecreaseAccuracy and get top
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  # Most informative variables
  importance <- rbind(importance, top_importance)  # Append to the importance table
  
# Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[seed] <- length(unique(noise$ASV))
unique_importance_count[seed]<- length(unique(importance$ASV))
}

# Summarize predictions
a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "L", "N", "P", "Q", "R")
colnames(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "N", "P", "Q", "R")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c / b

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")      
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'

# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%   # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe

```

```{r prediction_info}
always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

fall_Time0_prediction_info <- data.frame(
  Test = "fall day 0 soil",
  Train = "fall day 7 and 14 soil",
  Approach = "precence/absence",
  Accuracy = round(d, 4), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info , fall_Time0_prediction_info)
```

```{r confusion_matrix}
df <- as.data.frame(as.table(a))
Var1 <- unique(metaData_fall_Time0$place)

# Create a data frame with Var1, Var2 filled with "Q", and Freq filled with 0
df2 <- data.frame(
  Var1 = Var1,             # First column
  Var2 = rep("L", length(Var1)),  # Second column with "Q"
  Freq = rep(0, length(Var1))     # Third column with 0s
)
df <- rbind(df, df2)
df$Var2 <- as.character(df$Var2)
idx <- order(df$Var2)
df <- df[idx,]
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix_fall_prediction_on_fabric <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 22),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_fall_prediction_on_timepoint0_precence_absence.png", width = 1800, height = 1000)
print(conf_matrix_fall_prediction_on_fabric)
dev.off()
```

```{r Prob_distribution_by_location}
prob_plot <- ggplot(prob_long, aes(x = predicted_place, y = Probability, color = place)) +
  geom_point(position = position_jitter(width = 0.1), size = 2) +
  facet_wrap(~ place, scales = "free_x") +
  labs(title = "Dot Plot of Probabilities by Sample and Class",
       x = "Sample",
       y = "Probability") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
    text = element_text(size = 20),  # Global text size
    axis.title = element_text(size = 20),  # Axis titles
    axis.text = element_text(size = 20),  # Axis text
    legend.title = element_text(size = 20),  # Legend title
    legend.text = element_text(size = 20)   # Legend text
  )
png("results/Probabilities_by_location_ASV_fall_time0_precense_absense.png", width = 1800, height = 1000)
print(prob_plot)
dev.off()
```


```{r Prob_distribution}
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]

# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]


# Add an identifier column to each data frame
prob_long_cor$Prediction <- "Correct"
prob_long_wrong$Prediction <- "Incorrect"

# Combine the data frames
prob_long_combined <- rbind(prob_long_cor, prob_long_wrong)

prob_distribution <- ggplot(prob_long_combined, aes(x = Probability, fill = Prediction)) +
  geom_histogram(alpha = 0.5, bins = 200) +  # Use alpha for transparency
  labs(title = "Distribution of Probabilities (Correct vs Incorrect Predictions)",
       x = "Probability",
       y = "Density") +
  scale_fill_manual(values = c("Correct" = "skyblue", "Incorrect" = "salmon")) +
  theme_minimal()
png("results/probability_distribution_week0_precense_absense.png", width = 1800, height = 1000)
print(prob_distribution)
dev.off()
```

```{r unique_importance}
Importance_ASV_fall_Time0_precense_absense.tbl <- importance

```

# Seasonal

## Prediction on summer with training with fall

```{r subseting_table}
common_columns <- intersect(colnames(tab_rar_summer_ABCF), colnames(tab_rar_fall_ABCF))

# Subset both tables to only the common columns
tab_rar_summer_ABCF_subset <- tab_rar_summer_ABCF[, common_columns]
tab_rar_fall_ABCF_subset <- tab_rar_fall_ABCF[, common_columns]
```

```{r predict_pilot}
# Train and test data
x_train <- tab_rar_fall_ABCF_subset
y_train <- as.factor(metaData_fall_ABCF$place)  # target for training
x_test <- tab_rar_summer_ABCF_subset
y_test <- as.factor(metaData_summer_ABCF$place) 

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()  
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

# Loop over seeds from 1 to 10
for (seed in 1:5) {
  set.seed(seed)  # Set the seed

  # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  
                           mtry = floor(ncol(x_train)/6),  
                           num.threads = parallel::detectCores() - 1, 
                           importance = TRUE)
# Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)
  
prob <- predict(rf_model, x_test, type = "prob") # finding probability scores
prob_list[[seed]] <- prob  #saving the prob scores for each sample

#saving which samples were incorrectly predicted and what they were predicted as
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs here
      true_label = y_test[incorrect_idx], 
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
# Extract variable importance from the rf_model 
  importance_matrix <- as.data.frame(rf_model$importance) # Importance score for each ASV
  importance_matrix$ASV <-row.names(importance_matrix)
  
  # Sort by MeanDecreaseAccuracy and get top 
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  
  importance <- rbind(importance, top_importance)  # save the top scoring ASVs
  
  # Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[i] <- length(unique(noise$ASV))
unique_importance_count[i]<- length(unique(importance$ASV))
}

a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "F")
colnames(a) <- c("A", "C", "F")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c/b
print("Result matrix for fall sample prediction:")
print(a)
print(paste('Total nb of predicted samples:', b, sep=" "))
print(paste('Nb of correctly predicted samples', c, sep=" "))
print(paste('Proportion of correctly predicted samples', d, sep=" "))

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[seed]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")   
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'


# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%   # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe

```

```{r prediction_info}
always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

summer_prediction_fall_training_info <- data.frame(
  Test = "summer soil samples",
  Train = "fall soil samples",
  Approach = " ",
  Accuracy = round(d, 4), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info, summer_prediction_fall_training_info)
```

```{r conf_matrix}
df <- as.data.frame(as.table(a))
Var1 <- unique(metaData_fall_ABCF$place)

# Create a data frame with Var1, Var2 filled with "Q", and Freq filled with 0
df2 <- data.frame(
  Var1 = Var1,             # First column
  Var2 = rep("B", length(Var1)),  # Second column with "Q"
  Freq = rep(0, length(Var1))     # Third column with 0s
)
df <- rbind(df, df2)
df$Var2 <- as.character(df$Var2)
idx <- order(df$Var2)
df <- df[idx,]
head(df$Var2)
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix_TestSummer_TrainFall <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 30),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location", , title = "A") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_summer_prediction_with_fall_training.png", width = 1800, height = 1000)
print(conf_matrix_TestSummer_TrainFall)
dev.off()

```


```{r probability_by_location}
prob_plot <- ggplot(prob_long, aes(x = predicted_place, y = Probability, color = place)) +
  geom_point(position = position_jitter(width = 0.1), size = 2) +
  facet_wrap(~ place, scales = "free_x") +
  labs(title = "Dot Plot of Probabilities by Sample and Class",
       x = "Sample",
       y = "Probability") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
    text = element_text(size = 30),  # Global text size
    axis.title = element_text(size = 30),  # Axis titles
    axis.text = element_text(size = 30),  # Axis text
    legend.title = element_text(size = 30),  # Legend title
    legend.text = element_text(size = 30)   # Legend text
  )
png("results/Probabilities_by_location_ASV_predicting_summer_training_fall_plot.png", width = 1800, height = 1000)
print(prob_plot)
dev.off()
```

```{r importance}
Importance_ASV_summer_training_fall_all.tbl <- importance

```


### Precence abcence pred:

```{r predict_pilot}
tab_rar_summer_ABCF_subset[tab_rar_summer_ABCF_subset > 0] <- 1
tab_rar_fall_ABCF_subset[tab_rar_fall_ABCF_subset > 0] <- 1

# Train and test data
x_train <- tab_rar_fall_ABCF_subset
y_train <- as.factor(metaData_fall_ABCF$place)  # target for training
x_test <- tab_rar_summer_ABCF_subset
y_test <- as.factor(metaData_summer_ABCF$place) 

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()  
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

# Loop over seeds from 1 to 10
for (seed in 1:5) {
  set.seed(seed)  # Set the seed

  # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  
                           mtry = floor(ncol(x_train)/6),  
                           num.threads = parallel::detectCores() - 1, 
                           importance = TRUE)
# Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)
  
prob <- predict(rf_model, x_test, type = "prob") # finding probability scores
prob_list[[seed]] <- prob  #saving the prob scores for each sample

#saving which samples were incorrectly predicted and what they were predicted as
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs here
      true_label = y_test[incorrect_idx], 
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
# Extract variable importance from the rf_model 
  importance_matrix <- as.data.frame(rf_model$importance) # Importance score for each ASV
  importance_matrix$ASV <-row.names(importance_matrix)
  
  # Sort by MeanDecreaseAccuracy and get top 
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  
  importance <- rbind(importance, top_importance)  # save the top scoring ASVs
  
  # Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[i] <- length(unique(noise$ASV))
unique_importance_count[i]<- length(unique(importance$ASV))
}

a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "F")
colnames(a) <- c("A", "C", "F")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c/b

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[seed]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")   
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'


# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%   # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe

```

```{r prediction_info}
always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

summer_prediction_fall_training_pres_abs_info <- data.frame(
  Test = "summer soil samples",
  Train = "fall soil samples",
  Approach = "Presence/absence",
  Accuracy = round(d, 4), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info, summer_prediction_fall_training_pres_abs_info)
```


```{r conf_matrix}
df <- as.data.frame(as.table(a))
Var1 <- unique(metaData_fall_ABCF$place)

# Create a data frame with Var1, Var2 filled with "Q", and Freq filled with 0
df2 <- data.frame(
  Var1 = Var1,             # First column
  Var2 = rep("B", length(Var1)),  # Second column with "Q"
  Freq = rep(0, length(Var1))     # Third column with 0s
)
df <- rbind(df, df2)
df$Var2 <- as.character(df$Var2)
idx <- order(df$Var2)
df <- df[idx,]
head(df$Var2)
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 30),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location", , title = "A") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_summer_prediction_with_fall_training_presence_absence_plot.png", width = 1800, height = 1000)
print(conf_matrix)
dev.off()

```


```{r probability_by_location}
prob_plot <- ggplot(prob_long, aes(x = predicted_place, y = Probability, color = place)) +
  geom_point(position = position_jitter(width = 0.1), size = 2) +
  facet_wrap(~ place, scales = "free_x") +
  labs(title = "Dot Plot of Probabilities by Sample and Class",
       x = "Sample",
       y = "Probability") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
    text = element_text(size = 30),  # Global text size
    axis.title = element_text(size = 30),  # Axis titles
    axis.text = element_text(size = 30),  # Axis text
    legend.title = element_text(size = 30),  # Legend title
    legend.text = element_text(size = 30)   # Legend text
  )
png("results/Probabilities_by_location_ASV_predicting_summer_training_fall_Pres_abs_plot.png", width = 1800, height = 1000)
print(prob_plot)
dev.off()
```


```{r importance}
Importance_ASV_fall_training_with_summer_pres_abs.tbl <- importance
```

### Variable selection pred

```{r making_subsets}
asv_freq <- as.data.frame(table(Importance_ASV_summer_training_fall_all.tbl$ASV))
sorted_asv_freq <- asv_freq[order(-asv_freq$Freq), ]
```

```{r}
tab_rar_summer_ABCF_subset <- tab_rar_summer_ABCF[, common_columns]
tab_rar_fall_ABCF_subset <- tab_rar_fall_ABCF[, common_columns]
```

```{r}
# Finding the common entries
print(length(intersect(sorted_asv_freq$Var1, colnames(tab_rar_summer_ABCF_subset))))
```


```{r Finding_best_accuracy}
VS_across_season_summer_pred_info <- data.frame()
for (n_top_asvs in seq(10, 1000, by = 10)) {
  
  # Select top n ASVs
  top_ASVs <- head(sorted_asv_freq, n_top_asvs)
  selected_ASVs <- as.character(top_ASVs$Var1)
  

tab_summer <- tab_rar_summer_ABCF_subset[, colnames(tab_rar_summer_ABCF_subset) %in% selected_ASVs]
tab_fall <- tab_rar_fall_ABCF_subset[, colnames(tab_rar_fall_ABCF_subset) %in% selected_ASVs]


# Train and test data
x_train <- tab_fall
y_train <- as.factor(metaData_fall_ABCF$place)# target for training
x_test <- tab_summer
y_test <- as.factor(metaData_summer_ABCF$place) 

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()  
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)


# Loop over seeds from 1 to 10
for (seed in 1:5) {
  set.seed(seed)  # Set the seed

  # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  
                           mtry = floor(ncol(x_train)/6),  
                           num.threads = parallel::detectCores() - 1, 
                           importance = TRUE)
# Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)
  
prob <- predict(rf_model, x_test, type = "prob") # finding probability scores
prob_list[[seed]] <- prob  #saving the prob scores for each sample

#saving which samples were incorrectly predicted and what they were predicted as
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs here
      true_label = y_test[incorrect_idx], 
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
# Extract variable importance from the rf_model 
  importance_matrix <- as.data.frame(rf_model$importance) # Importance score for each ASV
  importance_matrix$ASV <-row.names(importance_matrix)
  
  # Sort by MeanDecreaseAccuracy and get top 
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  
  importance <- rbind(importance, top_importance)  # save the top scoring ASVs
  
  # Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[seed] <- length(unique(noise$ASV))
unique_importance_count[seed]<- length(unique(importance$ASV))
}


b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c/b

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")    
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'


# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%   # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe

VS_season_temp_info <- data.frame(
    Model = paste0("Top_", n_top_asvs, "_ASVs"),
    Accuracy = d,
    Features = ncol(x_train),
    Incorrect_Predicted_samples = paste0(incorrect_predictions$sample_id, collapse = ", "),
    Incorrect_Predicted_samples_freq = paste0(incorrect_predictions$freq, collapse = ", "),
    Predicted_as = paste0(incorrect_predictions$predicted_label, collapse = ", "),
    prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability)
  )
  
  VS_across_season_summer_pred_info <- rbind(VS_across_season_summer_pred_info, VS_season_temp_info)
}
```


```{r}
# Create the scatter plot
ggplot(VS_across_season_summer_pred_info, aes(x = Features, y = Accuracy)) +
  geom_point(size = 3, color = "steelblue") +
  geom_line(color = "steelblue", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Model Accuracy vs Number of Features",
    x = "Number of Features (ASVs)",
    y = "Prediction Accuracy" )
```

```{r predict_pilot}
top_ASVs <- head(sorted_asv_freq,  VS_across_season_summer_pred_info$Features[which.max(VS_across_season_summer_pred_info$Accuracy)])
selected_ASVs <- as.character(top_ASVs$Var1)

# Subset the feature table
tab_rar_summer_ABCF_subset <- tab_rar_summer_ABCF[,colnames(tab_rar_summer_ABCF) %in% selected_ASVs]
tab_rar_fall_ABCF_subset <- tab_rar_fall_ABCF[,colnames(tab_rar_fall_ABCF) %in% selected_ASVs]

# Train and test data
x_train <- tab_rar_fall_ABCF_subset
y_train <- as.factor(metaData_fall_ABCF$place)  # target for training
x_test <- tab_rar_summer_ABCF_subset
y_test <- as.factor(metaData_summer_ABCF$place) 

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()  
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

# Loop over seeds from 1 to 10
for (seed in 1:5) {
  set.seed(seed)  # Set the seed

  # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  
                           mtry = floor(ncol(x_train)/6),  
                           num.threads = parallel::detectCores() - 1, 
                           importance = TRUE)
# Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)
  
prob <- predict(rf_model, x_test, type = "prob") # finding probability scores
prob_list[[seed]] <- prob  #saving the prob scores for each sample

#saving which samples were incorrectly predicted and what they were predicted as
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs here
      true_label = y_test[incorrect_idx], 
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
# Extract variable importance from the rf_model 
  importance_matrix <- as.data.frame(rf_model$importance) # Importance score for each ASV
  importance_matrix$ASV <-row.names(importance_matrix)
  
  # Sort by MeanDecreaseAccuracy and get top 
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  
  importance <- rbind(importance, top_importance)  # save the top scoring ASVs
  
  # Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[i] <- length(unique(noise$ASV))
unique_importance_count[i]<- length(unique(importance$ASV))
}

a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "F")
colnames(a) <- c("A", "C", "F")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c/b

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[seed]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")   
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'


# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%   # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe

```

```{r prediction_info}
always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

summer_prediction_fall_training_pres_abs_info <- data.frame(
  Test = "summer soil samples",
  Train = "fall soil samples",
  Approach = "Feature selction",
  Accuracy = round(d, 4), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info, summer_prediction_fall_training_pres_abs_info)
```


```{r conf_matrix}
df <- as.data.frame(as.table(a))
Var1 <- unique(metaData_fall_ABCF$place)

# Create a data frame with Var1, Var2 filled with "Q", and Freq filled with 0
df2 <- data.frame(
  Var1 = Var1,             # First column
  Var2 = rep("B", length(Var1)),  # Second column with "Q"
  Freq = rep(0, length(Var1))     # Third column with 0s
)
df <- rbind(df, df2)
df$Var2 <- as.character(df$Var2)
idx <- order(df$Var2)
df <- df[idx,]
head(df$Var2)
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 30),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location", , title = "A") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_summer_prediction_with_fall_training_VS_plot.png", width = 1800, height = 1000)
print(conf_matrix)
dev.off()

```


## Predicting fall samples while training with summer
```{r predict_pilot}
# Subset both tables to only the common columns
tab_rar_summer_ABCF_subset <- tab_rar_summer_ABCF[, common_columns]
tab_rar_fall_ABCF_subset <- tab_rar_fall_ABCF[, common_columns]

# Train and test data
x_train <- tab_rar_summer_ABCF_subset
y_train <-   as.factor(metaData_summer_ABCF$place)# target for training
x_test <- tab_rar_fall_ABCF_subset
y_test <- as.factor(metaData_fall_ABCF$place) 

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()  
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

# Loop over seeds from 1 to 10
for (seed in 1:5) {
  set.seed(seed)  # Set the seed

  # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  
                           mtry = floor(ncol(x_train)/6),  
                           num.threads = parallel::detectCores() - 1, 
                           importance = TRUE)
# Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)
  
prob <- predict(rf_model, x_test, type = "prob") # finding probability scores
prob_list[[seed]] <- prob  #saving the prob scores for each sample

#saving which samples were incorrectly predicted and what they were predicted as
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs here
      true_label = y_test[incorrect_idx], 
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
# Extract variable importance from the rf_model 
  importance_matrix <- as.data.frame(rf_model$importance) # Importance score for each ASV
  importance_matrix$ASV <-row.names(importance_matrix)
  
  # Sort by MeanDecreaseAccuracy and get top 
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  
  importance <- rbind(importance, top_importance)  # save the top scoring ASVs
  
  # Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[seed] <- length(unique(noise$ASV))
unique_importance_count[seed]<- length(unique(importance$ASV))
}

a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "F")
colnames(a) <- c("A","B","C", "F")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c/b
print("Result matrix for fall sample prediction:")
print(a)
print(paste('Total nb of predicted samples:', b, sep=" "))
print(paste('Nb of correctly predicted samples', c, sep=" "))
print(paste('Proportion of correctly predicted samples', d, sep=" "))

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'


# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%   # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe
```


```{r prediction_info}
# Importance
Importance_ASV_fall_classification_with_summer_training.tbl <- unique(importance)

always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

fall_prediction_summer_training_info <- data.frame(
  Test = "fall soil samples",
  Train = "summer soil samples",
  Approach = " ",
  Accuracy = round(d, 4), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info, fall_prediction_summer_training_info)
```

```{r conf_matrix}
df <- as.data.frame(as.table(a))
# Calculate the row-wise sums
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix_TestFall_TrainSummer <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 30),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location", , title = "B") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_fall_prediction_with_summer_training.png", width = 1800, height = 1000)
print(conf_matrix_TestFall_TrainSummer)
dev.off()
```

```{r}
# Figure 5

Cross_seasonal_conf_matrixes <- conf_matrix_TestSummer_TrainFall + conf_matrix_TestFall_TrainSummer + plot_layout(axes = "collect")
png("results/Cross_seasonal_conf_matrixes.png", width = 900, height = 500)
print(Cross_seasonal_conf_matrixes)
dev.off()
```


### Predicting fall samples while training with summer presence/absence
```{r predict_pilot}
tab_rar_summer_ABCF_subset[tab_rar_summer_ABCF_subset > 0] <- 1
tab_rar_fall_ABCF_subset[tab_rar_fall_ABCF_subset > 0] <- 1


# Train and test data
x_train <- tab_rar_summer_ABCF_subset
y_train <-   as.factor(metaData_summer_ABCF$place)# target for training
x_test <- tab_rar_fall_ABCF_subset
y_test <- as.factor(metaData_fall_ABCF$place) 

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()  
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

# Loop over seeds from 1 to 10
for (seed in 1:5) {
  set.seed(seed)  # Set the seed

  # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  
                           mtry = floor(ncol(x_train)/6),  
                           num.threads = parallel::detectCores() - 1, 
                           importance = TRUE)
# Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)
  
prob <- predict(rf_model, x_test, type = "prob") # finding probability scores
prob_list[[seed]] <- prob  #saving the prob scores for each sample

#saving which samples were incorrectly predicted and what they were predicted as
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs here
      true_label = y_test[incorrect_idx], 
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
# Extract variable importance from the rf_model 
  importance_matrix <- as.data.frame(rf_model$importance) # Importance score for each ASV
  importance_matrix$ASV <-row.names(importance_matrix)
  
  # Sort by MeanDecreaseAccuracy and get top 
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  
  importance <- rbind(importance, top_importance)  # save the top scoring ASVs
  
  # Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[seed] <- length(unique(noise$ASV))
unique_importance_count[seed]<- length(unique(importance$ASV))
}

a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "F")
colnames(a) <- c("A","B","C", "F")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c/b
print("Result matrix for fall sample prediction:")
print(a)
print(paste('Total nb of predicted samples:', b, sep=" "))
print(paste('Nb of correctly predicted samples', c, sep=" "))
print(paste('Proportion of correctly predicted samples', d, sep=" "))

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'


# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%   # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe
```


```{r prediction_info}
# Importance
Importance_ASV_fall_classification_with_summer_training.tbl <- unique(importance)

always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

fall_prediction_summer_training_info <- data.frame(
  Test = "fall soil samples",
  Train = "summer soil samples",
  Approach = "presence/absence ",
  Accuracy = round(d, 4), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info, fall_prediction_summer_training_info)
```

```{r conf_matrix}
df <- as.data.frame(as.table(a))
# Calculate the row-wise sums
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix_prediction_on_summer_tFALL <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 30),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location", , title = "B") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_fall_prediction_with_summer_training_pres_abs.png", width = 1800, height = 1000)
print(conf_matrix_prediction_on_summer_tFALL)
dev.off()
```

### Predicting fall while training with summer - Variable selection
```{r making_subsets}
asv_freq <- as.data.frame(table(Importance_ASV_fall_classification_with_summer_training.tbl$ASV))
sorted_asv_freq <- asv_freq[order(-asv_freq$Freq), ]
```

```{r}
tab_rar_summer_ABCF_subset <- tab_rar_summer_ABCF[, common_columns]
tab_rar_fall_ABCF_subset <- tab_rar_fall_ABCF[, common_columns]
```

```{r}
# Finding the common entries
print(length(intersect(sorted_asv_freq$Var1, colnames(tab_rar_summer_ABCF_subset))))
```


```{r predict_fall}
VS_across_season_fall_pred_info <- data.frame()
for (n_top_asvs in seq(10, 1000, by = 10)) {
  
  # Select top n ASVs
  top_ASVs <- head(sorted_asv_freq, n_top_asvs)
  selected_ASVs <- as.character(top_ASVs$Var1)
  

tab_summer <- tab_rar_summer_ABCF_subset[, colnames(tab_rar_summer_ABCF_subset) %in% selected_ASVs]
tab_fall <- tab_rar_fall_ABCF_subset[, colnames(tab_rar_fall_ABCF_subset) %in% selected_ASVs]


# Train and test data
x_train <- tab_summer
y_train <- as.factor(metaData_summer_ABCF$place)# target for training
x_test <- tab_fall
y_test <- as.factor(metaData_fall_ABCF$place) 

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()  
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)


# Loop over seeds from 1 to 10
for (seed in 1:5) {
  set.seed(seed)  # Set the seed

  # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  
                           mtry = floor(ncol(x_train)/6),  
                           num.threads = parallel::detectCores() - 1, 
                           importance = TRUE)
# Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)
  
prob <- predict(rf_model, x_test, type = "prob") # finding probability scores
prob_list[[seed]] <- prob  #saving the prob scores for each sample

#saving which samples were incorrectly predicted and what they were predicted as
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs here
      true_label = y_test[incorrect_idx], 
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
# Extract variable importance from the rf_model 
  importance_matrix <- as.data.frame(rf_model$importance) # Importance score for each ASV
  importance_matrix$ASV <-row.names(importance_matrix)
  
  # Sort by MeanDecreaseAccuracy and get top 
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  
  importance <- rbind(importance, top_importance)  # save the top scoring ASVs
  
  # Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[seed] <- length(unique(noise$ASV))
unique_importance_count[seed]<- length(unique(importance$ASV))
}


b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c/b

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")    
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'


# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%   # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe

VS_season_temp_info <- data.frame(
    Model = paste0("Top_", n_top_asvs, "_ASVs"),
    Accuracy = d,
    Features = ncol(x_train),
    Incorrect_Predicted_samples = paste0(incorrect_predictions$sample_id, collapse = ", "),
    Incorrect_Predicted_samples_freq = paste0(incorrect_predictions$freq, collapse = ", "),
    Predicted_as = paste0(incorrect_predictions$predicted_label, collapse = ", "),
    prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability)
  )
  
  VS_across_season_fall_pred_info <- rbind(VS_across_season_fall_pred_info, VS_season_temp_info)
}
```


```{r}
# Create the scatter plot
ggplot(VS_across_season_fall_pred_info, aes(x = Features, y = Accuracy)) +
  geom_point(size = 3, color = "steelblue") +
  geom_line(color = "steelblue", linetype = "dashed") +
  theme_minimal() +
  labs(
    title = "Model Accuracy vs Number of Features",
    x = "Number of Features (ASVs)",
    y = "Prediction Accuracy" )
```

```{r predict_pilot}
# Subset both tables to only the common columns
top_ASVs <- head(sorted_asv_freq,  VS_across_season_fall_pred_info$Features[which.max(VS_across_season_fall_pred_info$Accuracy)])
selected_ASVs <- as.character(top_ASVs$Var1)

# Subset the feature table
tab_rar_summer_ABCF_subset <- tab_rar_summer_ABCF[,colnames(tab_rar_summer_ABCF) %in% selected_ASVs]
tab_rar_fall_ABCF_subset <- tab_rar_fall_ABCF[,colnames(tab_rar_fall_ABCF) %in% selected_ASVs]

# Train and test data
x_train <- tab_rar_summer_ABCF_subset
y_train <-   as.factor(metaData_summer_ABCF$place)# target for training
x_test <- tab_rar_fall_ABCF_subset
y_test <- as.factor(metaData_fall_ABCF$place) 

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()  
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

# Loop over seeds from 1 to 10
for (seed in 1:5) {
  set.seed(seed)  # Set the seed

  # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  
                           mtry = floor(ncol(x_train)/6),  
                           num.threads = parallel::detectCores() - 1, 
                           importance = TRUE)
# Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)
  
prob <- predict(rf_model, x_test, type = "prob") # finding probability scores
prob_list[[seed]] <- prob  #saving the prob scores for each sample

#saving which samples were incorrectly predicted and what they were predicted as
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs here
      true_label = y_test[incorrect_idx], 
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
# Extract variable importance from the rf_model 
  importance_matrix <- as.data.frame(rf_model$importance) # Importance score for each ASV
  importance_matrix$ASV <-row.names(importance_matrix)
  
  # Sort by MeanDecreaseAccuracy and get top 
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  
  importance <- rbind(importance, top_importance)  # save the top scoring ASVs
  
  # Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[seed] <- length(unique(noise$ASV))
unique_importance_count[seed]<- length(unique(importance$ASV))
}

a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "F")
colnames(a) <- c("A","B","C", "F")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c/b
print("Result matrix for fall sample prediction:")
print(a)
print(paste('Total nb of predicted samples:', b, sep=" "))
print(paste('Nb of correctly predicted samples', c, sep=" "))
print(paste('Proportion of correctly predicted samples', d, sep=" "))

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'


# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%   # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe
```


```{r prediction_info}
# Importance
Importance_ASV_fall_classification_with_summer_training.tbl <- unique(importance)

always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")


fall_prediction_summer_training_info <- data.frame(
  Test = "fall soil samples",
  Train = "summer soil samples",
  Approach = "Feature selection",
  Accuracy = round(d, 4), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info, fall_prediction_summer_training_info)
```

```{r conf_matrix}
df <- as.data.frame(as.table(a))
# Calculate the row-wise sums
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix_prediction_on_summer_tFALL <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 30),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location", , title = "B") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_fall_prediction_with_summer_training_VS.png", width = 1800, height = 1000)
print(conf_matrix_prediction_on_summer_tFALL)
dev.off()
```



# Fabric prediction

## Fabric summer
```{r}
common_columns <- intersect(colnames(tab_rar_summer_mF), colnames(tab_rar_summer_fabric))

# Subset both tables to only the common columns
tab_rar_summer_mF_subset <- tab_rar_summer_mF[, common_columns]
tab_rar_summer_fabric_subset <- tab_rar_summer_fabric[, common_columns]

# Train and test data
x_train <- tab_rar_summer_mF_subset  # feature matrix for training
y_train <- as.factor(metaData_summer_mF$place)  # target for training
x_test <- tab_rar_summer_fabric_subset  # feature matrix for testing
y_test <- as.factor(metaData_summer_fabric$place)  # target for testing (assuming first letter of row names)

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

for (seed in 1:5) {
  set.seed(seed)  # Set the seed
  
  # Train the Random Forest model
    # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  # Can adjust num.trees
                           mtry = floor(ncol(x_train)/6),  # Number of variables to consider at each split
                           importance = TRUE, num.threads = parallel::detectCores() - 1)
  
  # Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)

  # Get probabilities for the test set
  prob <- predict(rf_model, x_test, type = "prob")
  prob_list[[seed]] <- prob

  # Collect incorrect predictions
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs
      true_label = y_test[incorrect_idx],
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
  
  # Extract variable importance from the rf_model object directly
  importance_matrix <- as.data.frame(rf_model$importance)
  importance_matrix$ASV <- rownames(importance_matrix)  # Add ASV names
  importance_matrix$seed <- seed  # Add seed value
  
# Sort by MeanDecreaseAccuracy and get top
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  # Most informative variables
  importance <- rbind(importance, top_importance)  # Append to the importance table
  
# Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[seed] <- length(unique(noise$ASV))
unique_importance_count[seed]<- length(unique(importance$ASV))
}

# Summarize predictions
a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "E", "F", "G")
colnames(a) <- c("A", "B", "C", "E", "F", "G")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c / b


# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")    
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'

```

```{r prediction_info}

always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

summer_fabric_prediction_info <- data.frame(
  Test = "summer fabric samples",
  Train = "summer soil samples",
  Approach = "", 
  Accuracy = round(d, 4), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info , summer_fabric_prediction_info)
```

```{r confusion_matrix}
df <- as.data.frame(as.table(a))
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix_summer_prediction_on_fabric <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 22),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_summer_prediction_on_fabric.png", width = 1800, height = 1000)
print(conf_matrix_summer_prediction_on_fabric)
dev.off()
```

```{r Prob_distribution_by_location}
prob_plot <- ggplot(prob_long, aes(x = predicted_place, y = Probability, color = place)) +
  geom_point(position = position_jitter(width = 0.1), size = 2) +
  facet_wrap(~ place, scales = "free_x") +
  labs(title = "Dot Plot of Probabilities by Sample and Class",
       x = "Sample",
       y = "Probability") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
    text = element_text(size = 20),  # Global text size
    axis.title = element_text(size = 20),  # Axis titles
    axis.text = element_text(size = 20),  # Axis text
    legend.title = element_text(size = 20),  # Legend title
    legend.text = element_text(size = 20)   # Legend text
  )
png("results/Probabilities_by_location_ASV_summer_fabric_plot.png", width = 1800, height = 1000)
print(prob_plot)
dev.off()
```


```{r unique_importance}
Importance_ASV_summer_fabric.tbl <- importance
```



## Fabric Fall

```{r}
common_columns <- intersect(colnames(tab_rar_fall_mF), colnames(tab_rar_fall_fabric))

# Subset both tables to only the common columns
tab_rar_fall_mF_subset <- tab_rar_fall_mF[, common_columns]
tab_rar_fall_fabric_subset <- tab_rar_fall_fabric[, common_columns]

# Train and test data
x_train <- tab_rar_fall_mF_subset  # feature matrix for training
y_train <- as.factor(metaData_fall_mF$place)  # target for training
x_test <- tab_rar_fall_fabric_subset  # feature matrix for testing
y_test <- as.factor(metaData_fall_fabric$place)  # target for testing (assuming first letter of row names)

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

for (seed in 1:5) {
  set.seed(seed)  # Set the seed
  
  # Train the Random Forest model
    # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  # Can adjust num.trees
                           mtry = floor(ncol(x_train)/6),  # Number of variables to consider at each split
                           importance = TRUE, num.threads = parallel::detectCores() - 1)
  
  # Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)

  # Get probabilities for the test set
  prob <- predict(rf_model, x_test, type = "prob")
  prob_list[[seed]] <- prob

  # Collect incorrect predictions
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs
      true_label = y_test[incorrect_idx],
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
  
  # Extract variable importance from the rf_model object directly
  importance_matrix <- as.data.frame(rf_model$importance)
  importance_matrix$ASV <- rownames(importance_matrix)  # Add ASV names
  importance_matrix$seed <- seed  # Add seed value
  
# Sort by MeanDecreaseAccuracy and get top
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  # Most informative variables
  importance <- rbind(importance, top_importance)  # Append to the importance table
  
# Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[seed] <- length(unique(noise$ASV))
unique_importance_count[seed]<- length(unique(importance$ASV))
}

# Summarize predictions
a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "L", "N", "P", "Q", "R")
colnames(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "L", "N", "P", "R")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c / b

# Print results
print("Result matrix for fall sample prediction:")
print(a)
print(paste('Total number of predicted samples:', b, sep=" "))
print(paste('Number of correctly predicted samples:', c, sep=" "))
print(paste('Proportion of correctly predicted samples:', d, sep=" "))

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")    
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'

# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%   # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe

```

```{r prediction_info}
always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

fall_fabric_prediction_info <- data.frame(
  Test = "fall fabric samples",
  Train = "fall soil samples",
  Approach = "", 
  Accuracy = round(d, 4), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info , fall_fabric_prediction_info)
```

```{r confusion_matrix}
df <- as.data.frame(as.table(a))

Var1 <- unique(metaData_fall_mF$place)

# Create a data frame with Var1, Var2 filled with "Q", and Freq filled with 0
df2 <- data.frame(
  Var1 = Var1,             # First column
  Var2 = rep("Q", length(Var1)),  # Second column with "Q"
  Freq = rep(0, length(Var1))     # Third column with 0s
)
df <- rbind(df, df2)
df$Var2 <- as.character(df$Var2)
idx <- order(df$Var2)
df <- df[idx,]

row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix_fall_prediction_on_fabric <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 22),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_fall_prediction_on_fabric.png", width = 1800, height = 1000)
print(conf_matrix_fall_prediction_on_fabric)
dev.off()
```

```{r Prob_distribution_by_location}
prob_plot <- ggplot(prob_long, aes(x = predicted_place, y = Probability, color = place)) +
  geom_point(position = position_jitter(width = 0.1), size = 2) +
  facet_wrap(~ place, scales = "free_x") +
  labs(title = "Dot Plot of Probabilities by Sample and Class",
       x = "Sample",
       y = "Probability") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
    text = element_text(size = 20),  # Global text size
    axis.title = element_text(size = 20),  # Axis titles
    axis.text = element_text(size = 20),  # Axis text
    legend.title = element_text(size = 20),  # Legend title
    legend.text = element_text(size = 20)   # Legend text
  )
png("results/Probabilities_by_location_ASV_fall_fabric_plot.png", width = 1800, height = 1000)
print(prob_plot)
dev.off()
```


```{r unique_importance}
Importance_ASV_fall_fabric.tbl <- importance
```



## prediction of fabric from fall with variables selected as important during fall soil prediction -finding highest accuracy


```{r}
asv_freq <- as.data.frame(table(importance_ASV_fall_soil_samples.tbl$ASV))
sorted_asv_freq <- asv_freq[order(-asv_freq$Freq), ]

VS_fabric_info <- data.frame()

for (n_top_asvs in seq(10, 1000, by = 10)) {
  
  # Select top n ASVs
  top_ASVs <- head(sorted_asv_freq, n_top_asvs)
  selected_ASVs <- as.character(top_ASVs$Var1)
  
  common_columns <- intersect(colnames(tab_rar_fall_mF), colnames(tab_rar_fall_fabric))

# Subset both tables to only the common columns
tab_rar_fall_mF_subset <- tab_rar_fall_mF[, common_columns]
tab_rar_fall_fabric_subset <- tab_rar_fall_fabric[, common_columns]

tab_rar_fall_mF_subset <- tab_rar_fall_mF_subset[, colnames(tab_rar_fall_mF_subset) %in% selected_ASVs]
tab_rar_fall_fabric_subset <- tab_rar_fall_fabric_subset[, colnames(tab_rar_fall_fabric_subset) %in% selected_ASVs]

# Train and test data
x_train <- tab_rar_fall_mF_subset  # feature matrix for training
y_train <- as.factor(metaData_fall_mF$place)  # target for training
x_test <- tab_rar_fall_fabric_subset  # feature matrix for testing
y_test <- as.factor(metaData_fall_fabric$place)  # target for testing (assuming first letter of row names)

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

for (seed in 1:5) {
  set.seed(seed)  # Set the seed
  
  # Train the Random Forest model
    # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  # Can adjust num.trees
                           mtry = floor(ncol(x_train)/6),  # Number of variables to consider at each split
                           importance = TRUE, num.threads = parallel::detectCores() - 1)
  
  # Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)

  # Get probabilities for the test set
  prob <- predict(rf_model, x_test, type = "prob")
  prob_list[[seed]] <- prob

  # Collect incorrect predictions
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs
      true_label = y_test[incorrect_idx],
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
  
  # Extract variable importance from the rf_model object directly
  importance_matrix <- as.data.frame(rf_model$importance)
  importance_matrix$ASV <- rownames(importance_matrix)  # Add ASV names
  importance_matrix$seed <- seed  # Add seed value
  
# Sort by MeanDecreaseAccuracy and get top
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  # Most informative variables
  importance <- rbind(importance, top_importance)  # Append to the importance table
  
# Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[seed] <- length(unique(noise$ASV))
unique_importance_count[seed]<- length(unique(importance$ASV))
}

# Summarize predictions

b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c / b


# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")    
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'


# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%   # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe



  VS_fabric_temp_info <- data.frame(
    Model = paste0("Top_", n_top_asvs, "_ASVs"),
    Accuracy = d,
    Features = ncol(x_train),
    Incorrect_Predicted_samples = paste0(incorrect_predictions$sample_id, collapse = ", "),
    Incorrect_Predicted_samples_freq = paste0(incorrect_predictions$freq, collapse = ", "),
    Predicted_as = paste0(incorrect_predictions$predicted_label, collapse = ", "),
    prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability)
  )
  
  VS_fabric_info <- rbind(VS_fabric_info, VS_fabric_temp_info)
}
```

```{r}

# Create the scatter plot with a horizontal line at y = 0.876923076923077
ggplot(VS_fabric_info, aes(x = Features, y = Accuracy)) +
  geom_point(size = 3, color = "steelblue") +
  geom_line(color = "steelblue", linetype = "dashed") +
  geom_hline(yintercept = 0.876923076923077, color = "red", linetype = "solid") + # Add this line
  theme_minimal() +
  labs(
    title = "Model Accuracy vs Number of Features",
    x = "Number of Features (ASVs)",
    y = "Prediction Accuracy"
  )

```


## improving fabric fall pred


```{r}
common_columns <- intersect(colnames(tab_rar_fall_mF), colnames(tab_rar_fall_fabric))

# Subset both tables to only the common columns
tab_rar_fall_mF_subset <- tab_rar_fall_mF[, common_columns]
tab_rar_fall_fabric_subset <- tab_rar_fall_fabric[, common_columns]

tab_rar_fall_mF_subset <- tab_rar_fall_mF_subset[, colnames(tab_rar_fall_mF_subset) %in% selected_ASVs_fall_soil]
tab_rar_fall_fabric_subset <- tab_rar_fall_fabric_subset[, colnames(tab_rar_fall_fabric_subset) %in% selected_ASVs_fall_soil]

# Train and test data
x_train <- tab_rar_fall_mF_subset  # feature matrix for training
y_train <- as.factor(metaData_fall_mF$place)  # target for training
x_test <- tab_rar_fall_fabric_subset  # feature matrix for testing
y_test <- as.factor(metaData_fall_fabric$place)  # target for testing (assuming first letter of row names)

# Initialize containers
y_predTotal <- data.frame()
prob_list <- list()
incorrect_predictions <- data.frame()
importance <- data.frame()  # Container for importance information
noise <- data.frame()  # Container for noise variables (negative MeanDecreaseAccuracy)
unique_noise_count <- numeric()
unique_importance_count <- numeric()

for (seed in 1:5) {
  set.seed(seed)  # Set the seed
  
  # Train the Random Forest model
    # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  # Can adjust num.trees
                           mtry = floor(ncol(x_train)/6),  # Number of variables to consider at each split
                           importance = TRUE, num.threads = parallel::detectCores() - 1)
  
  # Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)

  # Get probabilities for the test set
  prob <- predict(rf_model, x_test, type = "prob")
  prob_list[[seed]] <- prob

  # Collect incorrect predictions
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs
      true_label = y_test[incorrect_idx],
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
  
  # Extract variable importance from the rf_model object directly
  importance_matrix <- as.data.frame(rf_model$importance)
  importance_matrix$ASV <- rownames(importance_matrix)  # Add ASV names
  importance_matrix$seed <- seed  # Add seed value
  
# Sort by MeanDecreaseAccuracy and get top
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  # Most informative variables
  importance <- rbind(importance, top_importance)  # Append to the importance table
  
# Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[seed] <- length(unique(noise$ASV))
unique_importance_count[seed]<- length(unique(importance$ASV))
}

# Summarize predictions
a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "L", "N", "P", "Q", "R")
colnames(a) <- c("A", "B", "C", "F", "H", "I", "J","K", "L", "N", "P", "R")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c / b
print(a)
# Print results
print(paste('Proportion of correctly predicted samples:', d, sep=" "))

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")    
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'


# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%   # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe
```

```{r prediction_info}
always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

fall_fabric_prediction_VS_info <- data.frame(
  Test = "fall fabric samples",
  Train = "fall soil samples",
  Approach = "Feature selection",
  Accuracy = round(d, 4), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)
prediction_info <- rbind(prediction_info , fall_fabric_prediction_VS_info)
```

```{r confusion_matrix}
df <- as.data.frame(as.table(a))
Var1 <- unique(metaData_fall_mF$place)

# Create a data frame with Var1, Var2 filled with "Q", and Freq filled with 0
df2 <- data.frame(
  Var1 = Var1,             # First column
  Var2 = rep("Q", length(Var1)),  # Second column with "Q"
  Freq = rep(0, length(Var1))     # Third column with 0s
)
df <- rbind(df, df2)
df$Var2 <- as.character(df$Var2)
idx <- order(df$Var2)
df <- df[idx,]
head(df$Var2)

row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))

# Plot the confusion matrix with custom axis titles
conf_matrix_fall_prediction_on_fabric <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 22),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_fall_fabric_VS.png", width = 1800, height = 1000)
print(conf_matrix_fall_prediction_on_fabric)
dev.off()
```

```{r Prob_distribution_by_location}
prob_plot <- ggplot(prob_long, aes(x = predicted_place, y = Probability, color = place)) +
  geom_point(position = position_jitter(width = 0.1), size = 2) +
  facet_wrap(~ place, scales = "free_x") +
  labs(title = "Dot Plot of Probabilities by Sample and Class",
       x = "Sample",
       y = "Probability") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
    text = element_text(size = 20),  # Global text size
    axis.title = element_text(size = 20),  # Axis titles
    axis.text = element_text(size = 20),  # Axis text
    legend.title = element_text(size = 20),  # Legend title
    legend.text = element_text(size = 20)   # Legend text
  )
png("results/Probabilities_by_location_ASV_fall_fabric_plot_VS.png", width = 1800, height = 1000)
print(prob_plot)
dev.off()
```

```{r investigating_noise}
unique_noise_df <- data.frame(
  seed = 1:20,  # Seed numbers from 1 to 20
  unique_count = unique_noise_count  # The unique noise counts
)
print(unique_noise_count)
```


# Variable selection based on Importance




## Investigating the important ASVs

```{r}

tab_rar_mF_fall_pres_abs <- tab_rar_fall_mF
tab_rar_mF_fall_pres_abs[tab_rar_mF_fall_pres_abs > 0] <- 1

# Calculate the sums of each column
precense <- as.data.frame(colSums(tab_rar_mF_fall_pres_abs))
colnames(precense) <- "nRows"
precense <- precense %>%
  mutate(Importance = ifelse(rownames(precense) %in% importance_ASV_fall_soil_samples.tbl$ASV, "Important", "Not Important"))


ggplot(precense, aes(x = Importance, y = nRows)) +
  geom_boxplot() +
  labs(title = "Box Plot of Mean Bray-Curtis by Variation",
       x = "Variation",
       y = "Mean Bray-Curtis") +
  theme_minimal()
mean(precense$nRows)
one <- filter(precense, Importance == "Important")
two <- filter(precense, Importance == "Not Important")

min(one$nRows)
wilcox.test(one$nRows, two$nRows, alternative = "two.sided")
min(one$nRows)
max(one$nRows)
abundance <- as.data.frame(colSums(tab_rar_fall_mF))
colnames(abundance) <- "nRows"
abundance <- abundance %>%
  mutate(Importance = ifelse(rownames(abundance) %in% importance_ASV_fall_soil_samples.tbl$ASV, "Important", "Not Important"))


ggplot(abundance, aes(x = Importance, y = nRows)) +
  geom_boxplot() +
  labs(title = "",
       x = "Asvs separated by if they are marked as important or not by random forest",
       y = "Scaled read counts") +
  theme_minimal()
one <- filter(abundance, Importance == "Important")
two <- filter(abundance, Importance == "Not Important")
wilcox.test(one$nRows, two$nRows, alternative = "two.sided")

```


```{r}

tab_rar_summer_pres_abs <- tab_rar_summer
tab_rar_summer_pres_abs[tab_rar_summer_pres_abs > 0] <- 1

# Calculate the sums of each column
precense <- as.data.frame(colSums(tab_rar_summer_pres_abs))
colnames(precense) <- "nRows"
precense <- precense %>%
  mutate(Importance = ifelse(rownames(precense) %in% unique(Importance_ASV_summer_all.tbl$ASV), "Important", "Not Important"))

unique(Importance_ASV_summer_all.tbl$ASV)
ggplot(precense, aes(x = Importance, y = nRows)) +
  geom_boxplot() +
  labs(title = "Box Plot of Mean Bray-Curtis by Variation",
       x = "Variation",
       y = "Mean Bray-Curtis") +
  theme_minimal()
mean(precense$nRows)
one <- filter(precense, Importance == "Important")
two <- filter(precense, Importance == "Not Important")

min(one$nRows)
wilcox.test(one$nRows, two$nRows, alternative = "two.sided")
min(one$nRows)
mean(one$nRows)

abundance <- as.data.frame(colSums(tab_rar_summer))
colnames(abundance) <- "nRows"
abundance <- abundance %>%
  mutate(Importance = ifelse(rownames(abundance) %in% Importance_ASV_summer_all.tbl$ASV, "Important", "Not Important"))


ggplot(abundance, aes(x = Importance, y = nRows)) +
  geom_boxplot() +
  labs(title = "",
       x = "Asvs separated by if they are marked as important or not by random forest",
       y = "Scaled read counts") +
  theme_minimal()
one <- filter(abundance, Importance == "Important")
two <- filter(abundance, Importance == "Not Important")
wilcox.test(one$nRows, two$nRows, alternative = "two.sided")
min(one$nRows)
mean(abundance$nRows)
```


```{r}
# Step 1: Get unique lengths
length_fall <- length(unique(importance_ASV_fall_soil_samples.tbl$ASV))
length_summer <- length(unique(Importance_ASV_summer_all.tbl$ASV))

# Step 2: Generate random subsets
set.seed(123)  # Setting a seed for reproducibility
random_subset_fall <- sample(colnames(tab_rar_fall_mF), length_fall, replace = FALSE)
random_subset_summer <- sample(colnames(tab_rar_summer), length_summer, replace = FALSE)

# Step 3: Calculate the intersections
original_intersect_length <- length(unique(intersect(importance_ASV_fall_soil_samples.tbl$ASV, unique(Importance_ASV_summer_all.tbl$ASV))))
random_intersect_length <- length(unique(intersect(random_subset_fall, random_subset_summer)))

# Step 4: Compare lengths
if (original_intersect_length > random_intersect_length) {
  print("The original intersection is longer than the random subset intersection.")
} else {
  print("The random subset intersection is longer than or equal to the original intersection.")
}

```

```{r}
set.seed(123)  # For reproducibility

# Number of permutations
n_permutations <- 1000
random_intersect_lengths <- numeric(n_permutations)

# Step 1: Get the original intersection length
original_intersect_length <- length(unique(intersect(
  importance_ASV_fall_soil_samples.tbl$ASV, 
  unique(Importance_ASV_summer_all.tbl$ASV)
)))

# Step 2: Generate random subsets and calculate intersection lengths
for (i in 1:n_permutations) {
  random_subset_fall <- sample(colnames(tab_rar_fall_mF), length_fall, replace = FALSE)
  random_subset_summer <- sample(colnames(tab_rar_summer), length_summer, replace = FALSE)
  
  # Calculate the intersection length
  random_intersect_lengths[i] <- length(unique(intersect(random_subset_fall, random_subset_summer)))
}
print(random_intersect_lengths)
# Step 3: Calculate the p-value
p_value <- mean(random_intersect_lengths >= original_intersect_length)

# Result output
cat("Original intersection length:", original_intersect_length, "\n")
cat("Mean of random intersect lengths:", mean(random_intersect_lengths), "\n")
cat("P-value:", p_value, "\n")

if (p_value < 0.05) {
    cat("The original intersection is significantly larger than expected by chance (p < 0.05).\n")
} else {
    cat("The original intersection is not significantly larger than expected by chance (p >= 0.05).\n")
}

```


```{r prediction_of_fall_soil_samples}
tab_rar_mF_fall_pres_abs_subset <- tab_rar_mF_fall_pres_abs[, colSums(tab_rar_mF_fall_pres_abs) > 3]

tab_rar_fall_mF_subset <- tab_rar_fall_mF[colnames(tab_rar_fall_mF) %in% colnames(tab_rar_mF_fall_pres_abs_subset)]

# Calculate the total abundance of each feature
total_abundance <- colSums(tab_rar_fall_mF)
head(total_abundance)
# Calculate the 90th percentile threshold
threshold <- quantile(total_abundance, 0.5)

# Subset the data frame to keep only features above the threshold
abundance_subset <- tab_rar_fall_mF[, total_abundance < threshold]


tab_rar_fall_mF_subset_1 <- tab_rar_fall_mF_subset[!(colnames(tab_rar_fall_mF_subset) %in% colnames(abundance_subset))]


# Define feature matrix (X) and target vector (y)
X <- tab_rar_fall_mF_subset_1
y <- as.factor(metaData_fall_mF$place)


folds_combined <- list()
for (i in 1:20) {
  set.seed(i)  # Set seed to the current value of i
  folds_combined <- c(folds_combined, createFolds(y, k = 5, list = TRUE, returnTrain = TRUE))
}


# Containers needed for collecting prediction info
y_predTotal <- data.frame()
prob_list <- list()  
incorrect_predictions <- data.frame() 
importance <- data.frame()  
noise <- data.frame()  
unique_noise_count <- numeric()
unique_importance_count <- numeric()

for (i in 1:25){ # Each sample is predicted 5 times
  set.seed(2)
  train_index <- unlist(folds_combined[[i]]) 
  x_train <- X[train_index, ]
  x_test <- X[-train_index, ]
  y_train <- y[train_index]
  y_test <- y[-train_index]
  
  # Train the Random Forest model
  rf_model <- randomForest(x = x_train, y = y_train, num.trees = 200,  
                           mtry = floor(ncol(x_train)/6),  
                           num.threads = parallel::detectCores() - 1, 
                           importance = TRUE)
# Predict on the test set
  y_pred <- predict(rf_model, x_test)
  names(y_pred) <- NULL
  y_predRes <- as.data.frame(cbind(y_pred, y_test))
  y_predTotal <- rbind(y_predTotal, y_predRes)
  
prob <- predict(rf_model, x_test, type = "prob") # finding probability scores
prob_list[[i]] <- prob  #saving the prob scores for each sample

#saving which samples were incorrectly predicted and what they were predicted as
  incorrect_idx <- which(y_pred != y_test)
  if (length(incorrect_idx) > 0) {
    incorrect_pred <- data.frame(
      sample_id = rownames(x_test)[incorrect_idx],  # Add sample IDs here
      true_label = y_test[incorrect_idx], 
      predicted_label = y_pred[incorrect_idx]
    )
    incorrect_predictions <- rbind(incorrect_predictions, incorrect_pred)
  }
# Extract variable importance from the rf_model 
  importance_matrix <- as.data.frame(rf_model$importance) # Importance score for each ASV
  importance_matrix$ASV <-row.names(importance_matrix)
  
  # Sort by MeanDecreaseAccuracy and get top 
  top_importance <- importance_matrix[order(-importance_matrix$MeanDecreaseAccuracy), ]  # Sort in descending order
  top_importance <- top_importance[1:min(Variables_selected, nrow(top_importance)), ]  
  importance <- rbind(importance, top_importance)  # save the top scoring ASVs
  
  # Identify variables with negative MeanDecreaseAccuracy
  noise_vars <- importance_matrix[importance_matrix$MeanDecreaseAccuracy < 0, ]
  noise <- rbind(noise, noise_vars)  # Append to the noise table

unique_noise_count[i] <- length(unique(noise$ASV))
unique_importance_count[i]<- length(unique(importance$ASV))
}

a <- table(y_predTotal$y_test, y_predTotal$y_pred)
row.names(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "L", "N", "P", "Q", "R")
colnames(a) <- c("A", "B", "C", "F", "H", "I", "J", "K", "L", "N", "P", "Q", "R")
b <- nrow(y_predTotal)
c <- sum(y_predTotal$y_pred == y_predTotal$y_test)
d <- c/b

# Prediction probabilities for the training data
prob <- do.call(rbind, lapply(1:length(prob_list), function(i) {
  fold_prob <- prob_list[[i]]
  data.frame(Sample = rownames(fold_prob), fold_prob, Fold = i)
}))

idx <- order(row.names(prob))
prob <- prob[idx,]
prob$place <- substring(row.names(prob),1,1)
prob_long <- melt(prob, id.vars = c("Sample", "place", "Fold"),  
                  variable.name = "predicted_place", 
                  value.name = "Probability")  
prob_long_cor <- prob_long[prob_long$place == prob_long$predicted_place, ]
prob_long_wrong <- prob_long[prob_long$place != prob_long$predicted_place, ]# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'

# inspecting incorrectly predicted samples
incorrect_predictions <- incorrect_predictions %>%
  group_by(sample_id) %>%   # Group by all unique columns
  summarize(freq = n()) %>%  # Count occurrences for each unique group and add 'freq' column
  ungroup()  # Ungroup to revert to standard dataframe

```

```{r prediction_info}
always_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq == 5], collapse = ", ")
intermittently_incorrect_ids <- paste(incorrect_predictions$sample_id[incorrect_predictions$freq < 5], collapse = ", ")

fall_soil_prediction_info <- data.frame(
  Test = "fall soil",
  Train = "fall soil",
  Approach = "cross-validation, -least abundant and least common",
  Accuracy = round(d, 4), 
  Features = ncol(x_train), 
  prob_dif = mean(prob_long_cor$Probability) - mean(prob_long_wrong$Probability),
  intermittently_incorrect = intermittently_incorrect_ids,
  always_incorrect = always_incorrect_ids)

prediction_info <- rbind(prediction_info , fall_soil_prediction_info)
```

```{r formating_confusion_matrix}
df <- as.data.frame(as.table(a))
# Calculate the row-wise sums
row_sums <- aggregate(Freq ~ Var1, data = df, sum)

# Join the summarized data back to the original dataframe
df <- df %>%
  left_join(row_sums, by = "Var1")

df <- df %>% 
  mutate(percent = round((Freq.x/Freq.y)*100)) %>%
  dplyr::select(Var1, Var2, percent)

df$Correct <- ifelse(df$percent == 0, "Zero",
                     ifelse(df$Var1 == df$Var2, "Correct", "Incorrect"))
# Reverse the order of the levels for the Y-axis (Var1) so the diagonal starts at the top-left
df$Var1 <- factor(df$Var1, levels = rev(levels(df$Var1)))


# Plot the confusion matrix with custom axis titles
conf_matrix_soil_fall <- ggplot(df, aes(x = Var2, y = Var1, fill = Correct)) +
  geom_tile(color = "black", size = 0.5) +
  geom_text(aes(label = percent), color = "black", size = 12) +  # Adjust text size
  scale_fill_manual(
    values = c("Correct" = "green", "Incorrect" = "red", "Zero" = "gray90"),
    breaks = c("Correct", "Incorrect")  # Show only Correct and Incorrect in the legend
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    # You can remove the element_blank() for axis titles
    axis.title.x = element_text(size = 22),  # Customize X-axis title font size
    axis.title.y = element_text(size = 22),   # Customize Y-axis title font size
  axis.text = element_text(size = 22),
  plot.title = element_text(size = 30),
    ) +
  labs(x = "Classified Location", y = "Sample Origin Location") +  # Add custom axis titles
  coord_fixed(ratio = 1)  + # Makes the cells square
  theme(legend.position = "none") 

png("results/conf_matrix_ASV_soil_fall_mostAbundant.png", width = 1800, height = 1000)
print(conf_matrix_soil_fall)
dev.off()
```

```{r Prob_distribution}
prob_long$Model <- rep("Soil_fall", length(prob_long$Sample))
prob_long_cor1 <- prob_long[prob_long$place == prob_long$predicted_place, ]

# Create prob_long_wrong for rows where 'place' does not match 'predicted_place'
prob_long_wrong1 <- prob_long[prob_long$place != prob_long$predicted_place, ]


# Add an identifier column to each data frame
prob_long_cor1$Prediction <- "Correct"
prob_long_wrong1$Prediction <- "Incorrect"

# Combine the data frames
prob_long_combined <- rbind(prob_long_cor1, prob_long_wrong1)

prob_distribution <- ggplot(prob_long_combined, aes(x = Probability, fill = Prediction)) +
  geom_histogram(alpha = 0.5, bins = 200) +  # Use alpha for transparency
  labs(title = "Distribution of Probabilities (Correct vs Incorrect Predictions)",
       x = "Probability",
       y = "Density") +
  scale_fill_manual(values = c("Correct" = "skyblue", "Incorrect" = "salmon")) +
  theme_minimal()
png("results/probability_distribution_ASV_fall_most_abundant_soil.png", width = 1800, height = 1000)
print(prob_distribution)
dev.off()
```

```{r Prob_distribution_by_location}
prob_plot <- ggplot(prob_long, aes(x = predicted_place, y = Probability, color = place)) +
  geom_point(position = position_jitter(width = 0.1), size = 2) +
  facet_wrap(~ place, scales = "free_x") +
  labs(title = "Dot Plot of Probabilities by Sample and Class",
       x = "Sample",
       y = "Probability") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), 
    text = element_text(size = 20),  # Global text size
    axis.title = element_text(size = 20),  # Axis titles
    axis.text = element_text(size = 20),  # Axis text
    legend.title = element_text(size = 20),  # Legend title
    legend.text = element_text(size = 20)   # Legend text
  )
png("results/Probabilities_by_location_ASV_fall_soil_most_abundant_and_commonASVs_plot.png", width = 1800, height = 1000)
print(prob_plot)
dev.off()
```
```{r importance_ASVs}
importance_ASV_fall_soil_samples.tbl <- importance
```



#Save prediction info

```{r}
write.table(prediction_info, prediction_info_file_out, sep = "\t")
```



```{r}
sessionInfo()
```





